---
title: "broccoli pre-MB permanova"
author: "Alexandra Alexiev"
date: "2024-10-02"
output: pdf_document
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
# knit options for whole doc. I will add in specific chunks when I don't want this to be the case
knitr::opts_chunk$set(echo = TRUE, # show code
                      eval = TRUE, # run chunks when knitting
                      include = TRUE, # include chunk input in final doc
                      warning = FALSE, # do not show warnings in final doc
                      message = FALSE, # do not show messages from code in final doc
                      collapse = TRUE, # when possible, do put multiple outputs in one block
                      dpi = 300, # fig resolution
                      fig.dim = c(9, 5), # the default figure dimensions
                      out.width = "98%", # the default figure output width
                      out.height = "98%", # the default figure output height 
                      cache = TRUE) # dont rerun chunks that haven't been changed

```

## libraries and directories

```{r libraries}
library(tibble)
library(dplyr)
library(vegan)
set.seed(3)

```

```{r set working dir, eval=F}
setwd("/Users/alexieva/Documents/Projects/Analysis/broccoli project/02_pubAnalysis/Enterotyping/")

load("/Users/alexieva/Documents/Projects/Analysis/broccoli project/02_pubAnalysis/Enterotyping//broccolipreMBpermanova.RData")

```

## Input data
```{r input microbiome asv table, urine SFN data, and diet data}
## microbiome data
# taxa table
MB_taxtab <- read.csv("/Users/alexieva/Documents/Projects/Analysis/broccoli project/00_data/Microbiome_Flat/TaxTable_BSS.csv",
                      header = TRUE) 

# metadata for microbiome stuff
metadata <- read.csv("/Users/alexieva/Documents/Projects/Analysis/broccoli project/00_data/Microbiome_Flat/SampleData_BSS.csv",
                      header = TRUE) %>%
  dplyr::filter(time == 0) %>% # for now just Time 0 data
  mutate(broc_consum = case_when(veg == "broc" ~ "1",
                                 veg == "alf" ~ "0"))

# ASV table (ASV reads per sample)
MB_asvtab <- read.csv("/Users/alexieva/Documents/Projects/Analysis/broccoli project/00_data/Microbiome_Flat/ASVTable_BSS.csv",
                      header = TRUE) %>%
  dplyr::filter(sample %in% metadata$sample) # just time 0 data

# clean up subject ID names so they will match the metadata
MB_asvtab$sample <- sub("_[^_]+_[^_]+$", "", MB_asvtab$sample)

# rename the ASVs by their taxonomy (genus level)
MB_asvtab <-as.data.frame(t(column_to_rownames(MB_asvtab, "sample")))
MB_asvtab$Genus[which(rownames(MB_asvtab) == MB_taxtab$ASV)] <- MB_taxtab$Genus
MB_asvtab <- MB_asvtab %>%
  group_by(Genus) %>%
  summarise(across(where(is.numeric), sum)) %>%
  column_to_rownames("Genus") %>%
  t() %>%
  as.data.frame()


```

```{r SFN data}
## SFN data
# cumulative across all timepoints
# include all SFN types
## SFN data

Urine_SFN <- read.csv("/Users/alexieva/Documents/Projects/Analysis/broccoli project/00_data/SFN_Data/Urine_R_Format.csv",
                      header = TRUE) %>%
    group_by(subject_id) %>%
  summarise(sum_SFN = sum(SFN),
            sum_SFN_Cys = sum(SFN_Cys),
            sum_SFN_NAC = sum(SFN_NAC),
            sum_SFN_CG = sum(SFN_CG),
            sum_SFN_GSH = sum(SFN_GSH),
            sum_SFN_NIT = sum(SFN_NIT),
            sum_SFN_Tot = sum(SFN_Tot))

# make a mega metadata file
mega_meta <- inner_join(metadata, Urine_SFN, by = "subject_id")
mega_meta <- mega_meta[order(mega_meta$subject_id),] # this also takes out the alfalfa eaters


```

```{r diet data}
## diet data
# make a list of the "significant" diet factors we are interested in 
diet_compnts <- c("diet_Chol_mg", "diet_SatFat_g", "diet_Alc_g", "diet_TotSolFib_g", 
                  "diet_MonSac_g", "diet_Disacc_g", "diet_BetaCaro_mcg", "diet_Vit_B1_mg",
                  "diet_Vit_B6_mg", "diet_Fol.DFE_mcg_DFE", "diet_Mang_mg", "diet_Copp_mg", 
                  "diet_TotFib_g", "diet_OCarb_g", "diet_Iron_mg", "diet_TotInsolFib_g")
diet <- read.csv("/Users/alexieva/Documents/Projects/Analysis/broccoli project/00_data/Diet/New_Diet_Small.csv", header = TRUE) %>%
  dplyr::select(c("subject_id", all_of(diet_compnts)))


```


```{r determine and assign enterotype}
library(phyloseq)

# make per sample taxa plots (stacked bar plots)
# ASV table (ASV reads per sample)
MB_asvtab_2 <- read.csv("/Users/alexieva/Documents/Projects/Analysis/broccoli project/00_data/Microbiome_Flat/ASVTable_BSS.csv",
                      header = TRUE) %>%
  dplyr::filter(sample %in% metadata$sample) # just time 0 data
# clean up subject ID names so they will match the metadata
MB_asvtab_2$sample <- sub("_[^_]+_[^_]+$", "", MB_asvtab_2$sample)
MB_asvtab_2 <- column_to_rownames(MB_asvtab_2, "sample")
# transpose
trans_mbasvtab <- as.data.frame(t(MB_asvtab_2))
# make col into rownames in taxa table
MB_taxtab2 <- MB_taxtab %>%
  column_to_rownames("ASV") %>%
  rename_at('Kingdom', ~ 'Domain') %>%
  as.matrix()

# make phyloseq obj and plot
OTU <- otu_table(trans_mbasvtab, taxa_are_rows = TRUE)
TAX <- tax_table(MB_taxtab2)
physeq <- phyloseq(OTU, TAX)
TopNOTUs <- names(which(MB_taxtab2[,6] == "Prevotella_9" | MB_taxtab2[,6] == "Bacteroides" | MB_taxtab2[,6] == "Ruminococcus"))
ent10 <- prune_taxa(TopNOTUs, physeq)
plot_bar(ent10, fill = "Genus")

# there does seem to be basis for 2-3 enterotypes here



```


## permanova tests on whole data set (with alfalfa eaters)
- wanted to test permanova models with alfalfa eaters and no SFN so there is variation in dietary components like fiber
```{r read in metadata files with groups (from different RMD)}
# with alfalfa eaters, there are 2 groups here, add diet
meta_groups_alfbrocc <- read.table("/Users/alexieva/Documents/Projects/Analysis/broccoli project/01_explAnalysis/brocc_mediation_Rfiles/inputs/meta_groups_alfbrocc.txt",
                                   header = T, sep = "\t") %>%
  inner_join(diet, by = "subject_id")

# without alfalfa eaters (just brocc), there are 3 groups here, add diet
meta_groups_brocc <- read.table("/Users/alexieva/Documents/Projects/Analysis/broccoli project/01_explAnalysis/brocc_mediation_Rfiles/inputs/meta_groups_brocc.txt",
                                   header = T, sep = "\t") %>%
  inner_join(diet, by = "subject_id")

```

- Ordistep used to choose best model for dbRDA.
- This best model was then used in the PERMANOVA to get significance.

```{r ordistep for dbRDA}
ALLmod0 <- capscale(MB_asvtab ~ 1, meta_groups_alfbrocc, distance = "bray")  # Model with intercept only
ALLmod1 <- capscale(MB_asvtab ~ . + .*., meta_groups_alfbrocc, distance = "bray")  # Model with all explanatory variables
## problem: model is overfitted

ordiALL <- ordistep(ALLmod0, scope = formula(ALLmod1)) # this determines what the best model is to run RDA on
ordiALL
# best model is MB_asvtab ~ groups_alf.cluster + diet_TotInsolFib_g + diet_OCarb_g + age
# + sample_issue + weight_lb + diet_TotInsolFib_g:age
# model inertia was 9.0628; Inertia is squared Bray distance


```

```{r dbRDA run with best model from ordistep}
# calculate RDA (distance based)
rda <- capscale(formula = MB_asvtab ~ groups_alf.cluster + diet_TotInsolFib_g + diet_OCarb_g + age
+ sample_issue + weight_lb + diet_TotInsolFib_g:age, 
                   data = meta_groups_alfbrocc, distance = "bray")
RsquareAdj(rda)
# 0.256 of variation explained by this model
summary(rda)

plot(rda)

```

```{r permanova best model}
dm_walf <- vegdist(MB_asvtab, method = "bray")
Perm <- adonis2(dm_walf ~ groups_alf.cluster + diet_TotInsolFib_g + diet_OCarb_g + age
+ sample_issue + weight_lb + diet_TotInsolFib_g:age, 
                            data = meta_groups_alfbrocc)
Perm

# write.table(Perm,
#             "Perm_w_alf.txt",
#             sep = "\t")


```

```{r rda graph}
# create files for graphing
smry_rda <- summary(rda)

# this is the coordinates for the points and metadata
meta_groups_filt <- meta_groups_alfbrocc %>%
  dplyr::select(c(subject_id, group, diet_TotInsolFib_g, diet_OCarb_g, veg, age, weight_lb, sample_issue))
pw_PC1  <- data.frame(smry_rda$sites[,1:2]) %>%  # these are the x, y coordinates for the sample points
  rownames_to_column("subject_id") %>%
  inner_join(meta_groups_filt, by = "subject_id") %>%
  column_to_rownames("subject_id")
mylims <- range(with(pw_PC1, c(CAP1, CAP2)))
pw_PC1$group <- as.factor(pw_PC1$group)

# the biplot scores are the correlations between your environmental variables and axes
pw_PC2  <- data.frame(smry_rda$biplot[1:6,])

# put plot together
rda_plot <- ggplot(pw_PC1, aes(x = CAP1, y = CAP2)) + 
  geom_point(aes(color = group), size = 2) +
  theme_classic() +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_vline(xintercept = 0, linetype = "dotted") +
  stat_ellipse(aes(group = veg, linetype = veg), show.legend = T) +
  geom_segment(data = pw_PC2, aes(x = 0, xend = CAP1, y = 0, yend = CAP2), 
               color = "black", arrow = arrow(length = unit(0.01, "npc"))) +
  geom_text(data = pw_PC2, 
            aes(x = CAP1, y = CAP2), label = row.names(pw_PC2), 
            color = "black", size = 4, hjust = 0, vjust = 1) +
  labs(x = paste0("CAP1 (",round(100*smry_rda$cont$importance[2, "CAP1"], digits = 2),"%)"),
       y = paste0("CAP2 (",round(100*smry_rda$cont$importance[2, "CAP2"], digits = 2),"%)")) +  
  scale_color_brewer(palette = "Set1",
                    name = "pre-treatment microbiome group") +
  theme(text = element_text(size = 20))
rda_plot

# ggsave("wALFpwrda_plot.tiff",
#        width = 12,
#        height = 7,
#        dpi = 300)


```


## permanova tests on only broccoli consumers
- removed alfalfa eaters and add sfn data
```{r filter so there are broccoli eaters only}
MB_asvtab_brocc <- MB_asvtab %>%
  rownames_to_column("subject_id") %>%
  dplyr::filter(subject_id %in% meta_groups_brocc$subject_id) %>%
  column_to_rownames("subject_id")


```

- Ordistep used to choose best model for dbRDA.
- This best model was then used in the PERMANOVA to get significance.

```{r broccoli ordistep for dbRDA}
meta_groups_brocc <- meta_groups_brocc %>%
  dplyr::select(-c(sample, treatment, sample_issue, time, veg, broc_consum, diet))
broccmod0 <- capscale(MB_asvtab_brocc ~ 1, meta_groups_brocc, distance = "bray")  # Model with intercept only
broccmod1 <- capscale(MB_asvtab_brocc ~ . + .*., meta_groups_brocc, distance = "bray")  # Model with all explanatory variables
ordibrocc <- ordistep(broccmod0, scope = formula(broccmod1)) # this determines what the best model is to run RDA on
ordibrocc
# best model is MB_asvtab_brocc ~ group + diet_TotSolFib_g + diet_OCarb_g + weight_lb +
# diet_TotSolFib_g:diet_OCarb_g
# model inertia was 4.6352; Inertia is squared Bray distance


```

```{r broccoli dbRDA run with best model from ordistep}
# calculate RDA (distance based)
rda_brocc <- capscale(formula = MB_asvtab_brocc ~ group + diet_TotSolFib_g + diet_OCarb_g + weight_lb +
diet_TotSolFib_g:diet_OCarb_g, data = meta_groups_brocc, 
                      distance = "bray")
RsquareAdj(rda_brocc)
# 0.374 of variation explained by this model
summary(rda_brocc)

plot(rda_brocc)

```

```{r broccoli permanova best model}
dm_brocc <- vegdist(MB_asvtab_brocc, method = "bray")
Perm_brocc <- adonis2(dm_brocc ~ group + diet_TotSolFib_g + diet_OCarb_g + weight_lb +
diet_TotSolFib_g:diet_OCarb_g, 
                            data = meta_groups_brocc)
Perm_brocc
p.adjust(Perm_brocc$`Pr(>F)`, "fdr")

# write.table(Perm_brocc,
#             "../outputs/Perm_brocc.txt",
#             sep = "\t")

```

```{r broccoli rda graph}
# create files for graphing
smry_rda_brocc <- summary(rda_brocc)

# this is the coordinates for the points and metadata
meta_groups_filt_brocc <- meta_groups_brocc %>%
  dplyr::select(c(subject_id, group, diet_OCarb_g, diet_TotSolFib_g, weight_lb))
brocc_PC1  <- data.frame(smry_rda_brocc$sites[,1:2]) %>%  # these are the x, y coordinates for the sample points
  rownames_to_column("subject_id") %>%
  inner_join(meta_groups_filt_brocc, by = "subject_id") %>%
  column_to_rownames("subject_id")
mylims <- range(with(brocc_PC1, c(CAP1, CAP2)))
brocc_PC1$group <- as.factor(brocc_PC1$group)

# the biplot scores are the correlations between your environmental variables and axes
brocc_PC2  <- data.frame(smry_rda_brocc$biplot[1:5,])

# put plot together
rda_plot_brocc <- ggplot(brocc_PC1, aes(x = CAP1, y = CAP2)) + 
  geom_point(aes(color = group), size = 2) +
  theme_classic() +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_vline(xintercept = 0, linetype = "dotted") +
  geom_segment(data = brocc_PC2[c(3:5),], aes(x = 0, xend = CAP1, y = 0, yend = CAP2), 
               color = "black", arrow = arrow(length = unit(0.01, "npc"))) +
  geom_text(data = brocc_PC2[c(3:5),], 
            aes(x = CAP1, y = CAP2), label = c("Total Soluble Fiber", "Other Carbs", "Weight (lb)"), 
            color = "black", size = 4, hjust = 1, vjust = 0.5) +
  labs(x = paste0("CAP1 (",round(100*smry_rda$cont$importance[2, "CAP1"], digits = 2),"%)"),
       y = paste0("CAP2 (",round(100*smry_rda$cont$importance[2, "CAP2"], digits = 2),"%)")) +  
  scale_color_brewer(palette = "Set1",
                    name = "pre-treatment\nmicrobiome group") +
  theme(text = element_text(size = 20))
rda_plot_brocc

# ggsave("../outputs/rda_plot_brocc_groupsinformula.png",
#        width = 10,
#        height = 7,
#        dpi = 300)


## simple graph
rda_plot_plain <- ggplot(brocc_PC1, aes(x = CAP1, y = CAP2)) + 
  geom_point(aes(color = group), size = 2) +
  theme_classic() +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_vline(xintercept = 0, linetype = "dotted") +
  scale_color_brewer(palette = "Set1",
                    name = "pre-treatment\nmicrobiome group") +
  theme(text = element_text(size = 20))
rda_plot_plain

# ggsave("../outputs/rda_plot_plain.png",
#        width = 10,
#        height = 7,
#        dpi = 300)

```

## redo above but without group - this is for the main manuscript
- broccoli only
- remove group since it might be "stealing" some of the variation in BC ordination space
- Sample issue removed; subject 40 is on the edge of the ordination space and was artificially stealing some variation as such but did not have any reason to be noted and isn't truly an outlier
```{r broccoli ordistep for dbRDA}
meta_nogroups_brocc <- meta_groups_brocc %>%
  dplyr::select(-c(group, ethnicity, relation, height_in, weight_lb)) %>% # removed colinear ones as well because they are creating overfitting in model
  column_to_rownames("subject_id")
broccmod0 <- capscale(MB_asvtab_brocc ~ 1, meta_nogroups_brocc, distance = "bray")  # Model with intercept only
broccmod1 <- capscale(MB_asvtab_brocc ~ . , meta_nogroups_brocc, distance = "bray")  # Model with all explanatory variables
ordibrocc <- ordistep(broccmod0, scope = formula(broccmod1), permutations = how(nperm = 499)) # this determines what the best model is to run RDA on
ordibrocc
# best model is diet_TotInsolFib_g + diet_OCarb_g
# model inertia was 4.6352; Inertia is squared Bray distance


```

```{r broccoli dbRDA run with best model from ordistep}
# calculate RDA (distance based) of first model
rda_brocc <- capscale(formula = MB_asvtab_brocc ~ diet_TotInsolFib_g + diet_OCarb_g, data = meta_nogroups_brocc, 
                      distance = "bray")
RsquareAdj(rda_brocc)
# 0.08 of variation explained by this model


```

```{r broccoli permanova best model}
Perm_brocc <- adonis2(formula = MB_asvtab_brocc ~ diet_TotInsolFib_g + diet_OCarb_g, 
                      data = meta_nogroups_brocc,
                      method = "bray",
                      by = "terms")
Perm_brocc
p.adjust(Perm_brocc$`Pr(>F)`, "fdr")

# write.table(Perm_brocc,
#             "../outputs/Perm_brocc_nogroups.txt",
#             sep = "\t")

```

```{r broccoli rda graph}
# create files for graphing
smry_rda_brocc <- summary(rda_brocc)

# this is the coordinates for the points and metadata
meta_groups_filt_brocc <- meta_groups_brocc %>%
  dplyr::select(c(subject_id, diet_TotInsolFib_g, diet_OCarb_g, group))
brocc_PC1  <- data.frame(smry_rda_brocc$sites[,1:2]) %>%  # these are the x, y coordinates for the sample points
  rownames_to_column("subject_id") %>%
  inner_join(meta_groups_filt_brocc, by = "subject_id") %>%
  column_to_rownames("subject_id")
mylims <- range(with(brocc_PC1, c(CAP1, CAP2)))
brocc_PC1$group <- as.factor(brocc_PC1$group)

# the biplot scores are the correlations between your environmental variables and axes
brocc_PC2  <- data.frame(smry_rda_brocc$biplot)

# create hulls
# hull_data <- brocc_PC1 %>%
#   group_by(group) %>%
#   slice(chull(CAP1, CAP2))

# put plot together
rda_plot_brocc <- ggplot(brocc_PC1, aes(x = CAP1, y = CAP2)) + 
  geom_point(aes(color = group), size = 2) +
  theme_classic() +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_vline(xintercept = 0, linetype = "dotted") +
  geom_segment(data = brocc_PC2, aes(x = 0, xend = CAP1, y = 0, yend = CAP2), 
               color = "black", arrow = arrow(length = unit(0.01, "npc"))) +
  geom_text(data = brocc_PC2, 
            aes(x = CAP1, y = CAP2), label = c("Total Insoluble Fiber (g)",
                                               "Other Carbs (g)"), 
            color = "black", size = 4, hjust = c(0,1), vjust = c(0, 0.5)) +
  # stat_ellipse(aes(linetype = group), position = "identity") +
  labs(x = paste0("CAP1 (",round(100*smry_rda$cont$importance[2, "CAP1"], digits = 2),"%)"),
       y = paste0("CAP2 (",round(100*smry_rda$cont$importance[2, "CAP2"], digits = 2),"%)")) +  
  scale_color_brewer(palette = "Set1",
                    name = "pre-treatment\nmicrobiome group") +
  theme(text = element_text(size = 20))
rda_plot_brocc

# ggsave("../outputs/rda_plot_brocc_V2.png",
#        width = 9,
#        height = 6,
#        dpi = 300)


```

## what could groups be correlated with?
```{r groups correlated with other vars}
# alfalfa and broccoli
meta_groups_alfbrocc$group_numb[meta_groups_alfbrocc$group == "A"] <- 1
meta_groups_alfbrocc$group_numb[meta_groups_alfbrocc$group == "B"] <- 2
names(meta_groups_alfbrocc)

lm_alfs <- lm(group_numb ~ sex + age + race + weight_lb + diet_condensed + cohort + ethnicity + height_in + sample_issue + diet_Chol_mg + diet_SatFat_g + diet_Alc_g + diet_TotSolFib_g + diet_MonSac_g + diet_Disacc_g + diet_BetaCaro_mcg + diet_Vit_B1_mg + diet_Vit_B6_mg + diet_Fol.DFE_mcg_DFE + diet_Mang_mg + diet_Copp_mg + diet_TotFib_g + diet_OCarb_g + diet_Iron_mg, 
              data = meta_groups_alfbrocc)
summary(lm_alfs)

# raceAsian                   1.587e+00  8.646e-01   1.835   0.0755 .
# raceDecline to state        2.244e+00  1.154e+00   1.944   0.0605 .
# raceMTOR                    1.823e+00  9.864e-01   1.848   0.0736 .
# raceOther                   2.438e+00  1.038e+00   2.349   0.0250 *
# raceWhite                   1.465e+00  8.537e-01   1.716   0.0956 .
# ethnicityNHL                9.577e-01  4.885e-01   1.961   0.0584 .
# diet_Disacc_g               9.206e+01  4.132e+01   2.228   0.0328 *
# diet_Mang_mg                3.301e+02  1.492e+02   2.212   0.0340 *

# broccoli only
meta_groups_brocc$group_numb[meta_groups_brocc$group == "A"] <- 1
meta_groups_brocc$group_numb[meta_groups_brocc$group == "B"] <- 2
meta_groups_brocc$group_numb[meta_groups_brocc$group == "C"] <- 3

lm_brocc <- lm(group_numb ~ sex + age + race + weight_lb + diet_condensed + cohort + ethnicity + height_in + diet_Chol_mg + diet_SatFat_g + diet_Alc_g + diet_TotSolFib_g + diet_MonSac_g + diet_Disacc_g + diet_BetaCaro_mcg + diet_Vit_B1_mg + diet_Vit_B6_mg + diet_Fol.DFE_mcg_DFE + diet_Mang_mg + diet_Copp_mg + diet_TotFib_g + diet_OCarb_g + diet_Iron_mg, 
              data = meta_groups_brocc)
sum <- summary(lm_brocc)

# age                         5.258e-02  1.609e-02   3.267  0.01710 *
# raceAsian                   4.083e+00  1.782e+00   2.292  0.06180 .
# raceMTOR                    4.792e+00  1.860e+00   2.577  0.04196 *
# raceWhite                   4.484e+00  1.752e+00   2.560  0.04293 *
# weight_lb                  -1.590e-02  7.742e-03  -2.053  0.08583 .
# diet_condensedPescetarian  -2.774e+00  7.440e-01  -3.729  0.00975 **
# diet_Alc_g                 -1.189e+02  5.092e+01  -2.336  0.05818 .
# diet_TotSolFib_g           -2.037e+03  6.975e+02  -2.920  0.02663 *
# diet_Disacc_g               2.432e+02  8.086e+01   3.008  0.02377 *
# diet_TotFib_g              -1.486e+02  6.761e+01  -2.198  0.07028 .
write.table(sum$coefficients, "../outputs/Pred2_grpLMs.txt")



```


## Permanova on each group
# group A
```{r broccoli ordistep for dbRDA group A}
meta_groupA_brocc <- meta_groups_brocc %>%
  dplyr::filter(group == "A") %>%
  dplyr::select(-c(group, group_numb, ethnicity, relation, height_in, weight_lb)) %>%
  column_to_rownames("subject_id")
MB_asvtab_broccA <- MB_asvtab_brocc %>%
  rownames_to_column("subject_id") %>%
  dplyr::filter(subject_id %in% rownames(meta_groupA_brocc)) %>%
  column_to_rownames("subject_id")
Abroccmod0 <- capscale(MB_asvtab_broccA ~ 1, meta_groupA_brocc, distance = "bray")  # Model with intercept only
Abroccmod1 <- capscale(MB_asvtab_broccA ~ . + .*., meta_groupA_brocc, distance = "bray")  # Model with all explanatory variables
ordibroccA <- ordistep(Abroccmod0, scope = formula(Abroccmod1)) # this determines what the best model is to run RDA on
ordibroccA
# best model is MB_asvtab_broccA ~ bmi + diet_Alc_g
# model inertia was 0.6843 (low, not great); Inertia is squared Bray distance


```

```{r broccoli dbRDA run with best model from ordistep group A}
# calculate RDA (distance based)
rda_broccA <- capscale(formula = MB_asvtab_broccA ~ bmi + diet_Alc_g, 
                      data = meta_groupA_brocc, 
                      distance = "bray")
RsquareAdj(rda_broccA)
# 0.17 of variation explained by this model
summary(rda_broccA)

plot(rda_broccA)

```

```{r broccoli permanova best model group A}
Perm_broccA <- adonis2(MB_asvtab_broccA ~ bmi + diet_Alc_g, 
                            data = meta_groupA_brocc,
                       method = "bray",
                       by = "terms")
Perm_broccA
p.adjust(Perm_broccA$`Pr(>F)`, method = "fdr")

# write.table(Perm_broccA,
#             "../outputs/Perm_broccA.txt",
#             sep = "\t")

```


# group B
```{r broccoli ordistep for dbRDA group A}
meta_groupB_brocc <- meta_groups_brocc %>%
  dplyr::filter(group == "B") %>%
  dplyr::select(-c(group, group_numb, ethnicity, relation, height_in, weight_lb)) %>%
  column_to_rownames("subject_id")
MB_asvtab_broccB <- MB_asvtab_brocc %>%
  rownames_to_column("subject_id") %>%
  dplyr::filter(subject_id %in% rownames(meta_groupB_brocc)) %>%
  column_to_rownames("subject_id")
Bbroccmod0 <- capscale(MB_asvtab_broccB ~ 1, meta_groupB_brocc, distance = "bray")  # Model with intercept only
Bbroccmod1 <- capscale(MB_asvtab_broccB ~ . + .*., meta_groupB_brocc, distance = "bray")  # Model with all explanatory variables
ordibroccB <- ordistep(Bbroccmod0, scope = formula(Bbroccmod1)) # this determines what the best model is to run RDA on
ordibroccB
# best model is MB_asvtab_broccB ~ race + diet_Vit_B6_mg + diet_Vit_B1_mg +
# race:diet_Vit_B1_mg
# model inertia was 1.91939; Inertia is squared Bray distance


```

```{r broccoli dbRDA run with best model from ordistep group A}
# calculate RDA (distance based)
rda_broccB <- capscale(formula = MB_asvtab_broccB ~ race + diet_Vit_B6_mg + diet_Vit_B1_mg +
race:diet_Vit_B1_mg, 
                      data = meta_groupB_brocc, 
                      distance = "bray")
RsquareAdj(rda_broccB)
# 0.3 of variation explained by this model
# race other and race white are collinear with vitamin B1
summary(rda_broccB)

plot(rda_broccB)

```

```{r broccoli permanova best model group A}
Perm_broccB <- adonis2(MB_asvtab_broccB ~ race + diet_Vit_B6_mg + diet_Vit_B1_mg +
race:diet_Vit_B1_mg, 
                            data = meta_groupB_brocc,
                       method = "bray",
                       by = "terms")
Perm_broccB
p.adjust(Perm_broccB$`Pr(>F)`, method = "fdr")

# write.table(Perm_broccB,
#             "../outputs/Perm_broccB.txt",
#             sep = "\t")

```

# group C 
```{r broccoli ordistep for dbRDA group A}
meta_groupC_brocc <- meta_groups_brocc %>%
  dplyr::filter(group == "C") %>%
  dplyr::select(-c(group, group_numb, ethnicity, relation, height_in, weight_lb)) %>%
  column_to_rownames("subject_id")
MB_asvtab_groupC <- MB_asvtab_brocc %>%
  rownames_to_column("subject_id") %>%
  dplyr::filter(subject_id %in% rownames(meta_groupC_brocc)) %>%
  column_to_rownames("subject_id")
Cbroccmod0 <- capscale(MB_asvtab_groupC ~ 1, meta_groupC_brocc, distance = "bray")  # Model with intercept only
Cbroccmod1 <- capscale(MB_asvtab_groupC ~ . + .*., meta_groupC_brocc, distance = "bray")  # Model with all explanatory variables
ordigroupC <- ordistep(Cbroccmod0, scope = formula(Cbroccmod1)) # this determines what the best model is to run RDA on
ordigroupC
# best model is MB_asvtab_groupC ~ diet_Fol.DFE_mcg_DFE
# model inertia was 0.4030 (low, not great); Inertia is squared Bray distance


```

```{r broccoli dbRDA run with best model from ordistep group A}
# calculate RDA (distance based)
rda_groupC <- capscale(formula = MB_asvtab_groupC ~ diet_Fol.DFE_mcg_DFE, 
                      data = meta_groupC_brocc, 
                      distance = "bray")
RsquareAdj(rda_groupC)
# 0.19 of variation explained by this model
summary(rda_groupC)

plot(rda_groupC)

```

```{r broccoli permanova best model group A}
Perm_groupC <- adonis2(MB_asvtab_groupC ~ diet_Fol.DFE_mcg_DFE, 
                            data = meta_groupC_brocc,
                       method = "bray",
                       by = "terms")
Perm_groupC
p.adjust(Perm_groupC$`Pr(>F)`, method = "fdr")

# write.table(Perm_groupC,
#             "../outputs/Perm_groupC.txt",
#             sep = "\t")

```

## save image chunk 
```{r save env, eval = F}
save.image("/Users/alexieva/Documents/Projects/Analysis/broccoli project/02_pubAnalysis/Enterotyping/broccolipreMBpermanova.RData")

```






