---
title: "brocc_mediation"
author: "Alexandra Alexiev"
date: "2024-07-09"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
# knit options for whole doc. I will add in specific chunks when I don't want this to be the case
knitr::opts_chunk$set(echo = TRUE, # show code
                      eval = TRUE, # run chunks when knitting
                      include = TRUE, # include chunk input in final doc
                      warning = FALSE, # do not show warnings in final doc
                      message = FALSE, # do not show messages from code in final doc
                      collapse = TRUE, # when possible, do put multiple outputs in one block
                      dpi = 300, # fig resolution
                      fig.dim = c(9, 5), # the default figure dimensions
                      out.width = "98%", # the default figure output width
                      out.height = "98%", # the default figure output height 
                      cache = TRUE) # dont rerun chunks that haven't been changed

```


## libraries and directories

```{r libraries}
library(dplyr)
library(tibble)
library(tidyr)
library(ggplot2)
library(mediation)
library(ggfortify)
library(ggpubr)
library(tweedie)
library(statmod)
library(AICcmodavg)
library(stringr)
set.seed(3)

```


```{r set working dir, eval=F}
setwd("/Users/alexieva/Documents/Projects/Analysis/broccoli project/02_pubAnalysis/Prediction1_mediation/")

load("/Users/alexieva/Documents/Projects/Analysis/broccoli project/02_pubAnalysis/Prediction1_mediation/brocc_mediation.RData")

```

## Input data

```{r input microbiome asv table, urine SFN data, and diet data}
## microbiome asv table; this is at time 0 as per previous exploratory data analysis
# only taxa that were previously identified as correlated with SFN and SFN-NIT, whose taxa names are in this SFN_taxa vector
SFN_taxa <- read.csv("/Users/alexieva/Documents/Projects/Analysis/broccoli project/01_explAnalysis/input_data/SFN_taxa.txt", header = T)

MB_asvtab <- read.table("/Users/alexieva/Documents/Projects/Analysis/broccoli project/01_explAnalysis/input_data/MBasvtab_brocc_cleanT0.txt", header = TRUE) %>%
  dplyr::select(SFN_taxa$x) %>%
  rownames_to_column("subject_id")

## SFN data
# using cumulative SFN, SFN-NIT, and SFN tot across all time points
# note that there are a few alfalfa eaters here that get filtered out in the process of making combined data frame of med_input
Urine_SFN <- read.csv("/Users/alexieva/Documents/Projects/Analysis/broccoli project/00_data/SFN_Data/Urine_R_Format.csv",
                      header = TRUE) %>%
  group_by(subject_id) %>%
  summarise(sum_SFN = sum(SFN),
            sum_SFN_Cys = sum(SFN_Cys),
            sum_SFN_NAC = sum(SFN_NAC),
            sum_SFN_CG = sum(SFN_CG),
            sum_SFN_GSH = sum(SFN_GSH),
            sum_SFN_NIT = sum(SFN_NIT),
            sum_SFN_Tot = sum(SFN_Tot))
  # %>% spread(key = "SFN_metabolites", value = "quantity")

## diet data
# make a list of the "significant" diet factors we are interested in 
diet_compnts <- c("diet_Chol_mg", "diet_SatFat_g", "diet_Alc_g", "diet_TotSolFib_g", 
                  "diet_MonSac_g", "diet_Disacc_g", "diet_BetaCaro_mcg", "diet_Vit_B1_mg",
                  "diet_Vit_B6_mg", "diet_Fol.DFE_mcg_DFE", "diet_Mang_mg", "diet_Copp_mg", 
                  "diet_TotFib_g", "diet_OCarb_g", "diet_Iron_mg", "diet_TotInsolFib_g")
diet <- read.csv("/Users/alexieva/Documents/Projects/Analysis/broccoli project/00_data/Diet/New_Diet_Small.csv", header = TRUE) %>%
  dplyr::select(c("subject_id", all_of(diet_compnts)))

# write.table(diet, "../input_data/dietData_clean.txt", row.names = F, quote = F)

```

```{r combine above separate data constructs into one with all info necessary for model}
# match data based on subject id
med_input <- MB_asvtab %>%
  inner_join(Urine_SFN, by = "subject_id") %>%
  inner_join(diet, by = "subject_id")

# write.table(med_input, "../input_data/combinedData_clean.txt", row.names = F, quote = F)

```

## LEFT OFF: rerunning this below with insol fiber data frames
## check distributions of variables
```{r check distrubutions of a couple ASVs we are interested in}
plot(density(MB_asvtab$Bifidobacterium))

plot(density(MB_asvtab$Lachnospiraceae.NK4A136.group))

plot(density(MB_asvtab$Ruminococcus))

plot(density(MB_asvtab$Roseburia))

plot(density(MB_asvtab$Blautia))
d <- density(MB_asvtab$Blautia)
quantile(MB_asvtab$Blautia)
d$x[which.max(d$y)] # use the max peak of the curve for treat.value in mediate() later
# this value is 1929.751

# most have right skewed density distributions with long right side tails, as is usual for ASV abundances, Blautia is at least approximately normal

```

```{r check distributions of SFN and SFN-NIT, just to get an idea}
plot(density(Urine_SFN$sum_SFN)) # kinda right skewed
plot(density(Urine_SFN$sum_SFN_NIT)) # this one is more normal but has a right tail
plot(density(Urine_SFN$sum_SFN_Tot)) # kinda right skewed

hist(Urine_SFN$sum_SFN)
hist(Urine_SFN$sum_SFN_NIT)
hist(Urine_SFN$sum_SFN_Tot)

```

```{r check distributions of total and soluble fiber}
plot(density(diet$diet_TotSolFib_g)) # very right skewed
plot(density(diet$diet_TotFib_g)) # more normal
plot(density(diet$diet_TotInsolFib_g)) # right skewed
plot(density(diet$diet_OCarb_g)) # more normal but still slightly right skewed
plot(density(diet$diet_Chol_mg)) # more normal but still slightly right skewed

```


## Run mediation analysis
- the hypothetical model we tested was whether certain ASVs (identified via partial correlation analysis; ASVs correlated to SFN and SFN-NIT) mediate the SFN response (SFN and SFN-NIT) to diet (based on several dietary components, particularly with interest in fiber)

try on one set of each X, Y, and M, then make into for loop
based on visual model here: https://library.virginia.edu/data/articles/introduction-to-mediation-analysis

X <- dietary components

Y <- SFN (3 types)

M <- ASVs

### First, check model 0 is true
- will use this output to also limit how many mediations I run later (i.e., only model 0 combinations of SFN and diet components will move forward to mediation analysis)

```{r choose a model for checking link between SFN/SFN-NIT and the different diet components}
# default lm
model.0.1 <- lm(sum_SFN ~ diet_TotFib_g, med_input)
# lm with log of SFN with pseudocount added
mod_log <- lm(log(sum_SFN+1) ~ diet_TotFib_g, med_input)
summary(mod_log)
plot(x = med_input$diet_TotFib_g, y = log(med_input$sum_SFN + 1))
# lm with polynomial
mod_logpoly <- lm(log(sum_SFN+1) ~ diet_TotFib_g + I(diet_TotFib_g^2) + I(diet_TotFib_g^3) + I(diet_TotFib_g^4), med_input)
summary(mod_logpoly)
plot(x = med_input$diet_TotFib_g, y = log(med_input$sum_SFN + 1))
pred <- predict(mod_logpoly)
ix <- sort(med_input$diet_TotFib_g, index.return = T)$ix
lines(med_input$diet_TotFib_g[ix], pred[ix], col='red', lwd = 2)
# gamma glm with pseudocount added to SFN
mod_gam <- glm((sum_SFN+1) ~ diet_TotFib_g, med_input, family = Gamma)
summary(mod_gam)
# inverse gaussian flm with pseudocount added to SFN
mod_invgaus <- glm((sum_SFN+1) ~ diet_TotFib_g, med_input, 
                   family = inverse.gaussian(link = "log"))
summary(mod_invgaus)
# tweedie
mod_tw <- glm((sum_SFN+1) ~ diet_TotFib_g, med_input,
              family = tweedie(var.power = 1, link.power = 0))
summary(mod_tw) # deviance super high
# negative binomial
mod_negbin <- glm.nb((sum_SFN+1) ~ diet_TotFib_g, med_input)
summary(mod_negbin)

# compare with aic
AIC(model.0.1, mod_log, mod_logpoly, mod_gam, mod_invgaus, mod_negbin)
# lm with log(SFN+1) is the best model, but the polynomial is close behind
# however, adjusted R^2 is higher in the logpoly model, so I should probably use that moving forward

```

```{r check if total fiber affects SFN production}
# model.0 <- lm(Y ~ X)
model.0.1 <- lm(log(sum_SFN+1) ~ diet_TotFib_g + I(diet_TotFib_g^2) + I(diet_TotFib_g^3) + I(diet_TotFib_g^4), med_input)
summary(model.0.1)
# every term is almost significant

# diagnostic plots
autoplot(model.0.1)
# Homoskedastic. Residuals are more or less normally distributed. Residuals are mostly linear.

# plot
plot(x = med_input$diet_TotFib_g, y = log(med_input$sum_SFN + 1))
pred <- predict(model.0.1)
ix <- sort(med_input$diet_TotFib_g, index.return = T)$ix
lines(med_input$diet_TotFib_g[ix], pred[ix], col='red', lwd = 2)


model.0.2 <- lm(log(sum_SFN_NIT+1) ~ diet_TotFib_g + I(diet_TotFib_g^2) + I(diet_TotFib_g^3) + I(diet_TotFib_g^4), med_input)
summary(model.0.2)
# no significance


```

```{r loop through all dietary components with each SFN}
list_sfn <- c("sum_SFN", "sum_SFN_Cys", "sum_SFN_NAC", "sum_SFN_CG", 
              "sum_SFN_GSH", "sum_SFN_NIT", "sum_SFN_Tot")
# and use diet_compnts as the list of dietary components
# make an output file
model.0_lms <- c()
pvals_model.0_lms <- c()

for (a in 1:length(list_sfn)) {
  for (e in 1:length(diet_compnts)) {
    model.0 <- lm(paste0("log(", list_sfn[a], "+1) ~ ", 
                        diet_compnts[e], "+ I(", diet_compnts[e], "^2) + I(", 
                        diet_compnts[e], "^3) + I(", diet_compnts[e], "^4)"), 
                        data = med_input)
    modsum <- summary(model.0)
    coeff_tab <- as.data.frame(modsum$coefficients)
    model.0_lms[[paste(list_sfn[a], "~", diet_compnts[e])]] <- coeff_tab
    pvals_model.0_lms[[paste(list_sfn[a], "~", diet_compnts[e])]][["diet"]] <- coeff_tab[2,4]
    pvals_model.0_lms[[paste(list_sfn[a], "~", diet_compnts[e])]][["diet2"]] <- coeff_tab[3,4]
    pvals_model.0_lms[[paste(list_sfn[a], "~", diet_compnts[e])]][["diet3"]] <- coeff_tab[4,4]
    pvals_model.0_lms[[paste(list_sfn[a], "~", diet_compnts[e])]][["diet4"]] <- coeff_tab[5,4]
    e <- e + 1
  }
  a <- a + 1
}

pvals_model.0_lms <- t(as.data.frame(pvals_model.0_lms))
# a few are significant with an alpha of 0.05

# want to save insoluble fiber to show coauthors and maybe reviewers
justinsols <- as.data.frame(pvals_model.0_lms) %>%
  filter(grepl("diet_TotInsolFib_g", rownames(pvals_model.0_lms)))
write.table(justinsols, "../outputs/insol_pvals.txt", quote = F)

hist(pvals_model.0_lms, breaks = 20)
abline(v = quantile(pvals_model.0_lms, 0.05))

# which ones are those?
sign_dietSFNcombos <- pvals_model.0_lms[which(pvals_model.0_lms <= 0.05), ]
sign_dietSFNcombos

# there are a ton of these that are significant

```

Pick model 0 as SFN and total dietary fiber to try code on

### Troubleshoot model Y and M that will go into mediation

```{r trying to troubleshoot model Y, eval=F}
## try to use a tweedie model
model.Y.1 <- glm(sum_SFN ~ diet_TotFib_g + Blautia, med_input, 
               family = tweedie(var.power = 1, link.power = 0))
# no warnings
summary(model.Y.1)
# its ok, no warnings now, but the deviance has gone up, no AIC, and more overdispersed

## try scaling sfn
# med_input$SFN_scaled <- round(((med_input$sum_SFN - min(med_input$sum_SFN)) / (max(med_input$sum_SFN) - min(med_input$sum_SFN)) * 100), digits = 0)
# plot(density(med_input$SFN_scaled))
# plot(density(med_input$sum_SFN))
# 
# model.Y.2 <- glm.nb(SFN_scaled ~ diet_TotFib_g + Blautia, med_input)
# summary(model.Y.2)
# plot(x = med_input$Blautia, y = med_input$SFN_scaled)
# # the same as unscaled so no point in scaling

## try log transforming
model.Y.3 <- glm.nb(log(sum_SFN+1) ~ diet_TotFib_g + Blautia, med_input)
summary(model.Y.3) # nothing
model.Y.3b <- glm.nb(sum_SFN ~ diet_TotFib_g + Blautia, med_input)

## try lm
model.Y.4 <- lm(log(sum_SFN+1) ~ diet_TotFib_g + log(Blautia+1), med_input)
summary(model.Y.4) # nothing

## inverse gaussian
model.Y.5 <- glm(formula = log(sum_SFN + 1) ~ diet_TotFib_g + Blautia, 
                family = inverse.gaussian(link = "log"), data = med_input)
summary(model.Y.5)


## inv gaussian wo log sfn
model.Y.5b <- glm(formula = sum_SFN ~ diet_TotFib_g + Blautia, 
                family = inverse.gaussian(link = "log"), data = med_input)

## use AIC to evaluate best option
AIC(model.Y.0, model.Y.2, model.Y.3, model.Y.3b, model.Y.4, model.Y.5, model.Y.5b)
AICtweedie( model.Y.1, dispersion = 1, k = 2, verbose = TRUE) # so really not a good fit at all for the data
# compare

# model Y4 has a negative adj R2 despite having the lowest AIC, so Y5 is the best choice here


```

```{r troubleshoot model.M, eval=F}
plot(med_input$quarts_totfib, med_input$f_Erysipelotrichaceae_ASV487)
plot(med_input$quarts_totfib, log(med_input$f_Erysipelotrichaceae_ASV487 + 1))

## basic lm
model.M.0 <- lm(f_Erysipelotrichaceae_ASV487 ~ diet_TotFib_g, med_input)
summary(model.M.0)

## glm negative binomial
model.M.1 <- glm.nb(f_Erysipelotrichaceae_ASV487 ~ diet_TotFib_g, med_input)
summary(model.M.1)

## log lm
model.M.2 <- lm(log(f_Erysipelotrichaceae_ASV487+1) ~ diet_TotFib_g, med_input)
summary(model.M.2)

## zero-infl neg bin
library(pscl)
model.M.3 <- zeroinfl(f_Erysipelotrichaceae_ASV487 ~ diet_TotFib_g, med_input, 
                      dist = "negbin")
summary(model.M.3)

## evaluate
AIC(model.M.0, model.M.1, model.M.2, model.M.3)
# model M2 is best (tried with a couple ASVs)

```

```{r run mediation code on test set of variables}
# add quartiles for X (diet_TotFib_g)
med_input <- med_input %>%
  mutate(quarts_totfib = ntile(diet_TotFib_g, 4),
         SFN_trans = log(sum_SFN+1),
         SFN_Tot_trans = log(sum_SFN_Tot+1),
         Blautia_trans = log(Blautia))
model.M <- lm(Blautia_trans ~ quarts_totfib, med_input)
model.Y <- lm(SFN_trans ~ quarts_totfib + Blautia, med_input)

results <- mediate(model.M, model.Y, treat = "quarts_totfib", mediator = "Blautia",
                   control.value = 1, treat.value = 4, # lowest and highest quartiles
                   boot = TRUE, sims = 500)
summary(results) # not great but at least now I have a way to implement loop code

```

## Implement loop to cycle through each relevant SFN-dietary combo and each ASV
- only use significant SFN-diet combos from above

```{r loop through all SFNs, dietary components, and ASVs}
# Use list_sfn and diet_compnts as the list of dietary components
# make SFN taxa a list
taxa_list <- as.character(SFN_taxa$x)
# make an output file
med_res <- c()
pvals_med <- c()
exp_res_med <- c()
list_combos <- as.data.frame(sign_dietSFNcombos) %>%
  rownames_to_column("comparison") %>%
  dplyr::filter(!grepl("diet2|diet3|diet4", comparison)) %>%
  mutate(comparison = str_remove(comparison, ".diet$"),
         sfn = stringr::str_split(comparison, "\\...", simplify = T)[, 1],
         diet = stringr::str_split(comparison, "\\...", simplify = T)[, 2],
         diet = replace(diet, diet == "diet_Fol", "diet_Fol.DFE_mcg_DFE")) %>%
  dplyr::select(-c("comparison", "sign_dietSFNcombos"))

for (a in 1:nrow(list_combos)) {
    for (f in 1:length(taxa_list)) {
      # add quartiles for each dietary component that will be used
      # transform ASV abundance and transform SFN abundance
      med_input <- med_input %>%
        dplyr::mutate(quarts = ntile(med_input[[list_combos[a,2]]], 4),
                      sfn_trans = log(med_input[[list_combos[a,1]]] + 1),
                      asv_trans = log(med_input[[taxa_list[f]]] + 1),
                      asv = med_input[[taxa_list[f]]])

      # set models
      model.M <- lm(asv_trans ~ quarts,
                    med_input)
      model.Y <- lm(sfn_trans ~ quarts + asv_trans,
                    med_input)
      
      # mediate
      results <- mediate(model.M, model.Y, 
                         treat = "quarts", 
                         mediator = "asv_trans",
                         control.value = 1, treat.value = 4, # lowest and highest quartiles
                         boot = TRUE, sims = 500)
      
      # save results
      exp_res_med[[paste(list_combos[a,1], taxa_list[f], list_combos[a,2])]] <- results
      modsum <- summary(results)
      pvals_med[[paste(list_combos[a,1], taxa_list[f], list_combos[a,2])]] <- list(ACME_pval = modsum$d.avg.p,
                                                                            ADE_pval = modsum$z.avg.p,
                                                                            prop_mediated_pval = modsum$n.avg.p)
      modsum$p.adj <- p.adjust(modsum$d.avg.p, method = "fdr")
      
      if (modsum$p.adj <= 0.2) {
        coeff_tab <- list(ACME_est = modsum$d.avg,
           ACME_pval = modsum$d.avg.p,
           ADE_est = modsum$z.avg,
           ADE_pval = modsum$z.avg.p,
           prop_mediated_est = modsum$n.avg,
           prop_mediated_pval = modsum$n.avg.p)

        med_res[[paste(list_combos[a,1], taxa_list[f], list_combos[a,2])]] <- coeff_tab

      } else { print(paste(list_combos[a,1], ",", taxa_list[f], ",",
                           list_combos[a,2], "are NOT significant"))}
      
      # counters
      f <- f + 1
  }
  a <- a + 1
}

```

```{r save the objects from having "or" between each p value in if statement above, eval = F}
# saveRDS(med_res, "mediation_result_notconservative.Rdata")
# saveRDS(pvals_med, "pvals_mediation_notconservative.Rdata")
# saveRDS(exp_res_med, "expanded_mediation_results_notconservative.Rdata")

# saveRDS(med_res, "mediation_result_signifsfndiets.Rdata")
# saveRDS(pvals_med, "pvals_mediation_signifsfndiets.Rdata")
# saveRDS(exp_res_med, "expanded_mediation_results_signifsfndiets.Rdata")

saveRDS(med_res, "mediation_result_signifcumSFNs.Rdata")
saveRDS(pvals_med, "pvals_mediation_signifcumSFNs.Rdata")
saveRDS(exp_res_med, "expanded_mediation_results_signifcumSFNs.Rdata")

```


## Examine results
```{r which SFN-taxa-diet combos are potential mediations}
signif <- do.call(rbind.data.frame, pvals_med[names(med_res)]) %>%
  rownames_to_column("names(med_res)")
signif

# write.table(signif, "../outputs/signif_meds_cumsfn.txt")

med_res

```

```{r for loop for all graphs}
# write a for loop with the above code to make graphs of all of them
medsignif_combos <- names(med_res) # 17 of these

# make output plot list
plot_list <- list()

# for loop to make graphs
for (i in 1:length(medsignif_combos)) {
  # first separate name to be three names
  name <- medsignif_combos[i]
  X <- strsplit(name, " ")[[1]][1]
  M <- strsplit(name, " ")[[1]][2]
  Y <- strsplit(name, " ")[[1]][3]
  
  # make a filtered data frame for ggplot to use
  regrplot <- med_input %>% 
    dplyr::select(all_of(M), all_of(X), all_of(Y))
  regrplot$trans_M <- log(regrplot[[M]]+1)
  regrplot$trans_X <- log(regrplot[[X]]+1)
  names(regrplot)[names(regrplot) == M] <- "M_vals"
  names(regrplot)[names(regrplot) == X] <- "X_vals"
  names(regrplot)[names(regrplot) == Y] <- "Y_vals"
  
  # make graph
  if(i == 1) {
    regrplot <- regrplot %>%
      dplyr::filter(regrplot$Y_vals != max(regrplot$Y_vals))
    }
  p <- ggplot(data = regrplot, aes(y = trans_X, x = trans_M)) +
    geom_point(aes(color = Y_vals)) +
    geom_smooth(se = TRUE, method = "lm") +
    labs(y = paste0(X, " quantity"),
       x = "ASV abundance",
       color = paste0("log transformed \n", Y, " quantity"), title = M) +
    theme(text = element_text(size = 14))
  
  # add to plot list
  plot_list[[i]] <- p
  # save
  ggsave(paste0("/Users/alexieva/Documents/Projects/Analysis/broccoli project/02_pubAnalysis/outputs/P1_medres/P1MedRes_plot_", i, ".png"),
         plot_list[[i]],
         height = 5, width = 7, dpi = 300)
  
  
}

g <- ggpubr::ggarrange(plotlist = plot_list,
          ncol = 5, nrow = 4,
          common.legend = T,
          legend = "bottom")

ggsave("/Users/alexieva/Documents/Projects/Analysis/broccoli project/02_pubAnalysis/outputs/P1MedRes_plotALL.png",
        g,
        height = 14, width = 18, dpi = 300)


```

```{r for loop for all graphs reversed SFN and diet}
# write a for loop with the above code to make graphs of all of them
medsignif_combos <- names(med_res) # 17 of these

# make output plot list
plot_list <- list()

# for loop to make graphs
for (i in 1:length(medsignif_combos)) {
  # first separate name to be three names
  name <- medsignif_combos[i]
  X <- strsplit(name, " ")[[1]][1]
  M <- strsplit(name, " ")[[1]][2]
  Y <- strsplit(name, " ")[[1]][3]
  
  # make a filtered data frame for ggplot to use
  regrplot <- med_input %>% 
    dplyr::select(all_of(M), all_of(X), all_of(Y))
  regrplot$trans_M <- log(regrplot[[M]]+1)
  regrplot$trans_X <- log(regrplot[[X]]+1)
  names(regrplot)[names(regrplot) == M] <- "M_vals"
  names(regrplot)[names(regrplot) == X] <- "X_vals"
  names(regrplot)[names(regrplot) == Y] <- "Y_vals"
  
  # make graph
  if(i == 1) {
    regrplot <- regrplot %>%
      dplyr::filter(regrplot$Y_vals != max(regrplot$Y_vals))
    }
  p <- ggplot(data = regrplot, aes(y = trans_M, x = Y_vals)) +
    geom_point(aes(color = trans_X)) +
    geom_smooth(se = TRUE, method = "lm") +
    labs(y = "ASV abundance",
       x = paste0("log transformed \n", Y, " quantity"),
       color = paste0(X, " quantity"), title = M) +
    theme(text = element_text(size = 14))
  
  # add to plot list
  plot_list[[i]] <- p
  # save
  ggsave(paste0("/Users/alexieva/Documents/Projects/Analysis/broccoli project/02_pubAnalysis/outputs/P1_medres/P1MedRes_plot_", i, ".png"),
         plot_list[[i]],
         height = 5, width = 7, dpi = 300)
  
  
}

g <- ggpubr::ggarrange(plotlist = plot_list,
          ncol = 5, nrow = 4,
          common.legend = T,
          legend = "bottom")

ggsave("/Users/alexieva/Documents/Projects/Analysis/broccoli project/02_pubAnalysis/outputs/P1MedRes_plotALL_reversed.png",
        g,
        height = 14, width = 18, dpi = 300)


```

## Re-run Collinsella result without Vit B1 outlier person
- the significant combo from above, Collinsella - vitamin B1 - SFN, has an outlier in the vit B1 data (one inidividual just has extremely high B1 for some reason)
- plan is to remove this outlier and rerun just that mediation and graph

```{r collinsella B1 SFN rerun filter outlier from input}
med_input_B1filt <- med_input %>%
  dplyr::filter(diet_Vit_B1_mg != max(diet_Vit_B1_mg))


```

```{r collinsella B1 SFN rerun mediation}
med_input_B1filt2 <- med_input_B1filt %>%
        dplyr::mutate(quarts = ntile(diet_Vit_B1_mg, 4),
                      asv_trans = log(Collinsella+1),
                      sfn_trans = log(sum_SFN+1))

model.M.B1 <- lm(asv_trans ~ quarts,
                    med_input_B1filt2)
summary(model.M)

model.Y.B1 <- lm(sfn_trans ~ quarts + asv_trans,
                    med_input_B1filt2)
summary(model.Y)

med_B1results <- mediate(model.M.B1, model.Y.B1, 
                         treat = "quarts", 
                         mediator = "asv_trans",
                         control.value = 1, treat.value = 4, # lowest and highest quartiles
                         boot = TRUE, sims = 500)


modsumB1 <- summary(med_B1results)
modsumB1

coeff_tab_medB1 <- list(ACME_est = modsumB1$d.avg,
           ACME_pval = p.adjust(modsumB1$d.avg.p, method = "fdr"),
           ADE_est = modsumB1$z.avg,
           ADE_pval = p.adjust(modsumB1$z.avg.p, method = "fdr"),
           prop_mediated_est = modsumB1$n.avg,
           prop_mediated_pval = p.adjust(modsumB1$n.avg.p, method = "fdr"))


```


## save image chunk 
```{r save env, eval = F}
save.image("/Users/alexieva/Documents/Projects/Analysis/broccoli project/02_pubAnalysis/Prediction1_mediation/brocc_mediation.RData")

```
