---
title: "broccoli random forest"
author: "Alexandra Alexiev"
date: "2024-11-18"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
# knit options for whole doc. I will add in specific chunks when I don't want this to be the case
knitr::opts_chunk$set(echo = TRUE, # show code
                      eval = TRUE, # run chunks when knitting
                      include = TRUE, # include chunk input in final doc
                      warning = FALSE, # do not show warnings in final doc
                      message = FALSE, # do not show messages from code in final doc
                      collapse = TRUE, # when possible, do put multiple outputs in one block
                      dpi = 300, # fig resolution
                      fig.dim = c(9, 5), # the default figure dimensions
                      out.width = "98%", # the default figure output width
                      out.height = "98%", # the default figure output height 
                      cache = TRUE) # dont rerun chunks that haven't been changed

```


## libraries and directories

```{r libraries}
library(dplyr)
library(tibble)
library(ggplot2)
library(randomForest)
set.seed(3)

```


```{r set working dir, eval=F}
setwd("/Users/alexieva/Documents/Projects/Analysis/broccoli project/02_pubAnalysis/RandomForest/")

load("/Users/alexieva/Documents/Projects/Analysis/broccoli project/02_pubAnalysis/RandomForest/brocc_randomforestSFNNIT.RData")

```

## Input files prep

We want an input file that has the "features" we want to test the importance of vs the samples. The features we want there are the microbial taxa followed by the SFN_NIT data.
```{r file input and clean up}
## input microbiome taxa table
MB_asvtab <- read.delim("/Users/alexieva/Documents/Projects/Analysis/broccoli project/01_explAnalysis/input_data/MBasvtab_brocc_cleanT0.txt") %>%
  rownames_to_column("subject_id")
# str(MB_asvtab)
# MB_asvtab[,2:112] <- log(MB_asvtab[,2:112]+1)

## input SFN_NIT data but combine with asv tab
SFN_NIT_tab <- read.delim("/Users/alexieva/Documents/Projects/Analysis/broccoli project/01_explAnalysis/input_data/Urine_SFN_cumul.txt")

# no NA's?
sum(!complete.cases(MB_asvtab))
sum(!complete.cases(SFN_NIT_tab))
# yes

# for now test with SFN_NIT_cumulative data
feat_tab <- MB_asvtab %>%
  inner_join(SFN_NIT_tab[c("subject_id", "sum_SFN_NIT")], by = "subject_id") %>%
  column_to_rownames("subject_id")

```

## graph the distribution of SFN-NIT and a test taxon
These were logged in the optimal model we used in the mediation analysis so we intend to do that here for the random forest regressions as well.
```{r SFN-NIT and test taxon graphs}
hist(feat_tab$sum_SFN_NIT) # non-normal
hist(feat_tab$Dialister) # also non-normal

hist(log(feat_tab$sum_SFN_NIT)) # more normal
hist(log(feat_tab$Dialister)) # also more normal

# let's just log the whole matrix (with pseudocount), as we did in the mediation analysis
feat_tabl <- log(feat_tab + 1)


```

## Split into train and validation set of data
```{r train and validation set split}
# Training Set : Validation Set = 70 : 30 (random)
train <- sample(nrow(feat_tabl), 0.7*nrow(feat_tabl), replace = FALSE)
TrainSet <- feat_tabl[train,]
ValidSet <- feat_tabl[-train,]
summary(TrainSet)
summary(ValidSet)


```

## create random forest model to test which taxa are associated with cumulative SFN_NIT
- default uses 500 trees
- regression model

```{r create default random forest model on training set}
# Create a Random Forest model with default parameters
model1 <- randomForest(sum_SFN_NIT ~ ., data = TrainSet, importance = TRUE)
model1
round(importance(model1), 2)

# Get variable importance from the model fit
ImpData <- as.data.frame(importance(model1))
ImpData$Var.Names <- row.names(ImpData)

ggplot(ImpData, aes(x=Var.Names, y=`%IncMSE`)) +
  geom_segment( aes(x=Var.Names, xend=Var.Names, y=0, yend=`%IncMSE`), color="skyblue") +
  geom_point(aes(size = IncNodePurity), color="blue", alpha=0.6) +
  theme_light() +
  coord_flip() +
  theme(
    legend.position="bottom",
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank()
  )
# since some of these are negative make sure later we use absolute valyue MSE greater than 1

# ggsave("import_graph_trainSFNNIT.png",
#        dpi = 300, height = 13,
#        width = 8)

# calculate RMSE of train data
sqrt(model1$mse[length(model1$mse)]) 
# 0.7410906
# want to be low; SFN-NIT logged ranges up to 5

# calculate RMSE of test data
predValues <- predict(model1, ValidSet)
sqrt(mean((ValidSet$sum_SFN_NIT - predValues)^2))
# 0.7835532

R2 <- 1 - (sum((ValidSet$sum_SFN_NIT-predValues)^2)/sum((ValidSet$sum_SFN_NIT-mean(ValidSet$sum_SFN_NIT))^2))
# -0.612301
# not a great fit for data? but also found an article that describes that it *could* 
# be negative because we used different sets of data to train and test so the sums of 
# squares are not exactly comparable; this is probably not the best metric for fit then
# https://towardsdatascience.com/explaining-negative-r-squared-17894ca26321/ 

# find number of trees that produced lowest test MSE
which.min(model1$mse) # 25

# RMSE of best model
sqrt(model1$mse[which.min(model1$mse)])
# 0.7045018; pretty good

#plot the test MSE by number of trees
plot(model1)

```

```{r run full model}
model2 <- randomForest(formula = sum_SFN_NIT ~ . , data = feat_tabl, importance = TRUE)
model2 # not much different than train model

# find number of trees that produced lowest test MSE
which.min(model2$mse)
# 126 trees best, but this changes every time I run it

# RMSE of best model
sqrt(model2$mse[which.min(model2$mse)])
# 0.7368217 is best mse

#plot the test MSE by number of trees
plot(model2)

# Get variable importance from the model fit
ImpData <- as.data.frame(importance(model2))
ImpData$Var.Names <- row.names(ImpData)

ggplot(ImpData, aes(x=Var.Names, y=`%IncMSE`)) +
  geom_segment( aes(x=Var.Names, xend=Var.Names, y=0, yend=`%IncMSE`), color="skyblue") +
  geom_point(aes(size = IncNodePurity), color="blue", alpha=0.6) +
  theme_light() +
  coord_flip() +
  theme(
    legend.position="bottom",
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank()
  )

# ggsave("import_graph.png",
#        dpi = 300, height = 13,
#        width = 8)


```

optimize model:
- ntreeTry: The number of trees to build.
- mtryStart: The starting number of predictor variables to consider at each split.
- stepFactor: The factor to increase by until the out-of-bag estimated error stops improving by a certain amount.
- improve: The amount that the out-of-bag error needs to improve by to keep increasing the step factor.
```{r optimize/tune the rf model}
model_tuned <- tuneRF(
               x=feat_tabl[,-112], #define predictor variables
               y=feat_tabl$sum_SFN_NIT, #define response variable
               ntreeTry=500,
               mtryStart=3, 
               stepFactor=1.5,
               improve=0.01,
               trace=FALSE #don't show real-time progress
               )
# 3 predictors (mtry) at each split will give best OOB

```

```{r rerun rf model with new information about best number of predictors}
# run model with best trees
model3 <- randomForest(formula = sum_SFN_NIT ~ . , data = feat_tabl, 
                       importance = TRUE, ntree = 500, mtry = 4)
model3 # not much different than train model

# RMSE of best model
sqrt(model3$mse[which.min(model3$mse)])
# 0.7108695
# compare to range of Dialister abundance for e.g.
range(feat_tabl$Dialister)
# so the error is not too bad

#plot the test MSE by number of trees
plot(model3)

# Get variable importance from the model fit
ImpData <- as.data.frame(importance(model3)) %>%
  rename_at('%IncMSE', ~'percIncMSE') %>%
  dplyr::filter(percIncMSE > 1)
ImpData$Var.Names <- row.names(ImpData)

mod3Impgraph <- ggplot(ImpData, aes(x=Var.Names, y=`percIncMSE`)) +
  geom_segment( aes(x=Var.Names, xend=Var.Names, y=0, yend=`percIncMSE`), color="skyblue") +
  geom_point(aes(size = IncNodePurity), color="blue", alpha=0.6) +
  theme_light() +
  coord_flip() +
  theme(legend.position="bottom",
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    text = element_text(size = 14)) +
  labs(y = "Percent increase in MSE", x = "", size = "Increase in Node Purity")
mod3Impgraph

# ggsave("import_graph_optimtestmod_SFNNIT.png",
#        dpi = 300, height = 9,
#        width = 8)


```

```{r calculate metrics of model fit}
# psuedo r squared
predValues <- predict(model3, ValidSet)
R2 <- 1 - (sum((ValidSet$sum_SFN_NIT-predValues)^2)/sum((ValidSet$sum_SFN_NIT-mean(ValidSet$sum_SFN_NIT))^2))
R2
# pretty ok
# 0.7452101


```

```{r graph and linear regression of dialister and SFN-NIT}
# list the taxa with an perincmse higher than 2 and node purity inc higher than 0.1
top_tax <- ImpData %>%
  dplyr::filter(percIncMSE > 2 & IncNodePurity > 0.1)
# note this uses logged tax abundances and SFN
toptax_graph <- feat_tabl %>%
  dplyr::select(c(rownames(top_tax), "sum_SFN_NIT"))

# stats
summary(lm(sum_SFN_NIT ~ Alistipes, data = toptax_graph)) # nope
summary(lm(sum_SFN_NIT ~ Dialister, data = toptax_graph)) # nope
summary(lm(sum_SFN_NIT ~ Lachnospiraceae.UCG.009, data = toptax_graph)) # almost
summary(lm(sum_SFN_NIT ~ Oscillospira, data = toptax_graph)) # yes
summary(lm(sum_SFN_NIT ~ f_NA_ASV156, data = toptax_graph)) # yes

# make a little table of these to add as annotations to graph later
lachno <- summary(lm(sum_SFN_NIT ~ Lachnospiraceae.UCG.009, data = toptax_graph)) # almost
osci <- summary(lm(sum_SFN_NIT ~ Oscillospira, data = toptax_graph)) # yes
NAasv <- summary(lm(sum_SFN_NIT ~ f_NA_ASV156, data = toptax_graph)) # yes
stat_tab <- data.frame(Taxon = c(rownames(lachno$coefficients)[2],
                                 rownames(osci$coefficients)[2],
                                 rownames(NAasv$coefficients)[2]),
                       Est = round(c(lachno$coefficients[2,1],
                                 osci$coefficients[2,1],
                                 NAasv$coefficients[2,1]), digits = 1),
                       Error = round(c(lachno$coefficients[2,2],
                                 osci$coefficients[2,2],
                                 NAasv$coefficients[2,2]), digits = 2),
                       pval = round(c(lachno$coefficients[2,4],
                                 osci$coefficients[2,4],
                                 NAasv$coefficients[2,4]), digits = 2))
# turn into text object for.graph
text_annot <- data.frame(label = c(paste0("Estimate = ", stat_tab$Est[1],
                                                " +/- ", stat_tab$Error[1], "\n",
                                                "p = ", stat_tab$pval[1]),
                                   paste0("Estimate = ", stat_tab$Est[2],
                                                " +/- ", stat_tab$Error[2], "\n",
                                                "p = ", stat_tab$pval[2]),
                                   paste0("Estimate = ", stat_tab$Est[3],
                                                " +/- ", stat_tab$Error[3], "\n",
                                                "p = ", stat_tab$pval[3])),
                         Taxon = stat_tab$Taxon)

# change format for graphing
toptax_graph_long <- toptax_graph %>%
  tidyr::gather(key = "Taxon", value = "abund", -sum_SFN_NIT)

# graph
taxnit_plot <- ggplot(data = toptax_graph_long,
                       aes(x = abund, y = sum_SFN_NIT)) +
  geom_point() +
  geom_smooth(method = 'lm', color = "black") +
  labs(x = "Log taxon abundance", y = "log cumulative SFN-nitrile in urine") +
  theme(text = element_text(size = 14)) +
  facet_wrap("Taxon")

taxnit_plot

# ggsave("TOP5NIT_lmplot.jpg", taxnit_plot,
#        dpi = 300, width = 7, height = 5)

# different version of graph
# change format for graphing
toptax_graph_long2 <- toptax_graph %>%
  dplyr::select(c("Oscillospira", "f_NA_ASV156", "Lachnospiraceae.UCG.009", "sum_SFN_NIT")) %>%
  tidyr::gather(key = "Taxon", value = "abund", -sum_SFN_NIT)
# graph
taxnit_plot2 <- ggplot(data = toptax_graph_long2,
                       aes(x = abund, y = sum_SFN_NIT)) +
  geom_point() +
  geom_smooth(method = 'lm', color = "black") +
  labs(x = "Log taxon abundance", y = "log cumulative SFN-nitrile in urine") +
  theme(text = element_text(size = 14)) +
  facet_wrap("Taxon") +
  geom_text(data = text_annot, mapping = aes(x = 2.5, y = 2, label = label))

taxnit_plot2

# ggsave("TOP3NIT_lmplot.jpg", taxnit_plot2,
#        dpi = 300, width = 9, height = 4)


```

```{r make a paneled figure of importance graph and regressions}
library(ggpubr)
implm <- ggarrange(mod3Impgraph, taxnit_plot2,
                   nrow = 2, ncol = 1,
                   heights = c(8, 4),
                   labels = c("A", "B"))
implm

# ggsave("imp_and_lm_rfplot.jpg", implm,
#        dpi = 300, width = 9, height = 12)


```

## random forest model to test whether the presence or absence of certain taxa is associated with cumulative SFN_NIT

```{r convert asv table into presence absence and add groups}
metadata <- read.delim("../input_data/metadata_wgroups_brocc.txt",
                       header = T, sep = "\t") %>%
  dplyr::select(c("subject_id", "group"))
MB_asvtab_pa <- column_to_rownames(MB_asvtab, "subject_id")
MB_asvtab_pa <- as.data.frame(ifelse(MB_asvtab_pa == 0, "A", "P")) %>%
  rownames_to_column("subject_id") %>%
  inner_join(SFN_NIT_tab[c("subject_id", "sum_SFN_NIT")], by = "subject_id") %>%
  column_to_rownames("subject_id")

## separate into each group
MBasvtab_pa_A <- dplyr::filter(MB_asvtab_pa %>% rownames_to_column("subject_id"), 
                               subject_id %in% metadata$subject_id[metadata$group == "A"])
MBasvtab_pa_B <- dplyr::filter(MB_asvtab_pa %>% rownames_to_column("subject_id"), 
                               subject_id %in% metadata$subject_id[metadata$group == "B"])
MBasvtab_pa_C <- dplyr::filter(MB_asvtab_pa %>% rownames_to_column("subject_id"), 
                               subject_id %in% metadata$subject_id[metadata$group == "C"])


```

```{r split microbiome group data into train and valid set}
# Training Set : Validation Set = 70 : 30 (random)
list <- list(MBasvtab_pa_A, MBasvtab_pa_B, MBasvtab_pa_C)

for (i in list) {
  train <- sample(nrow(i), 0.7*nrow(i), replace = FALSE)
  TrainSet <- i[train,]
  ValidSet <- i[-train,]
  
}

# for whole data
train <- sample(nrow(MB_asvtab_pa), 0.7*nrow(MB_asvtab_pa), replace = FALSE)
TrainSet <- MB_asvtab_pa[train,]
ValidSet <- MB_asvtab_pa[-train,]
summary(TrainSet)
summary(ValidSet)


```

```{r create default random forest model on training set}
# Create a Random Forest model with default parameters
model1 <- randomForest(sum_SFN_NIT ~ ., data = TrainSet, importance = TRUE, type = "class")
model1

R2 <- 1 - (sum((ValidSet$sum_SFN_NIT-predValues)^2)/sum((ValidSet$sum_SFN_NIT-mean(ValidSet$sum_SFN_NIT))^2))
# negative, not a great fit for data

# find number of trees that produced lowest test MSE
which.min(model1$mse)
# 500

# RMSE of best model
sqrt(model1$mse[which.min(model1$mse)])
# 27.4

```

```{r run full model}
model2 <- randomForest(formula = sum_SFN_NIT ~ . , data = MB_asvtab_pa, importance = TRUE)
model2 # not much different than train model

# find number of trees that produced lowest test MSE
which.min(model2$mse)
# 106 trees best

# RMSE of best model
sqrt(model1$mse[which.min(model2$mse)])
# 28 is best mse

#plot the test MSE by number of trees
plot(model2)


```

optimize model:
- ntreeTry: The number of trees to build.
- mtryStart: The starting number of predictor variables to consider at each split.
- stepFactor: The factor to increase by until the out-of-bag estimated error stops improving by a certain amount.
- improve: The amount that the out-of-bag error needs to improve by to keep increasing the step factor.
```{r optimize/tune the rf model}
model_tuned <- tuneRF(
               x=MB_asvtab_pa[,-112], #define predictor variables
               y=MB_asvtab_pa$sum_SFN_NIT, #define response variable
               ntreeTry=500,
               mtryStart=2, 
               stepFactor=1.5,
               improve=0.01,
               trace=FALSE #don't show real-time progress
               )
# 2 predictors at each split will give best OOB

```

```{r rerun rf model with new information about best number of predictors}
model3 <- randomForest(formula = sum_SFN_NIT ~ . , data = MB_asvtab_pa, importance = TRUE, mtry = 2)
model3 # not much different than train model

# find number of trees that produced lowest test MSE
which.min(model3$mse)
# 29 trees best, but this changes every time I run it

# RMSE of best model
sqrt(model3$mse[which.min(model3$mse)])
# 7.8 is best mse

# r squared
predValues <- predict(model3, ValidSet)
R2 <- 1 - (sum((ValidSet$sum_SFN_NIT-predValues)^2)/sum((ValidSet$sum_SFN_NIT-mean(ValidSet$sum_SFN_NIT))^2))
# pretty mid

#plot the test MSE by number of trees
plot(model3)

# Get variable importance from the model fit
ImpData <- as.data.frame(model3[["importance"]]) %>%
  rename_at('%IncMSE', ~'percIncMSE') %>%
  dplyr::filter(percIncMSE > 0)
ImpData$Var.Names <- row.names(ImpData)

ggplot(ImpData, aes(x=Var.Names, y=`percIncMSE`)) +
  geom_segment( aes(x=Var.Names, xend=Var.Names, y=0, yend=`percIncMSE`), color="skyblue") +
  geom_point(aes(size = IncNodePurity), color="blue", alpha=0.6) +
  theme_light() +
  coord_flip() +
  theme(
    legend.position="bottom",
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank()
  )

# ggsave("import_graph_optimtestmod_presabstaxa.png",
#        dpi = 300, height = 8,
#        width = 8)

# another verison of importance plot
varImpPlot(model3, 
           sort=FALSE, 
           main="Variable Importance Plot")


```

## random forest model to test whether microbiome grouping is more important than dietary params in determining SFN_NIT metrics in broccoli consumers
```{r input data with microbiome groups}
MB_wgroups <- read.delim("meta_groups_brocc.txt", header = T, sep = " ") %>%
  dplyr::select(-c("group_numb", "sample", "treatment", "relation")) %>%
  inner_join(SFN_NIT_tab[c("subject_id", "sum_SFN_NIT")], by = "subject_id") %>%
  column_to_rownames("subject_id")


```

```{r split microbiome group data into train and valid set}
# Training Set : Validation Set = 70 : 30 (random)
train <- sample(nrow(MB_wgroups), 0.7*nrow(MB_wgroups), replace = FALSE)
TrainSet <- MB_wgroups[train,]
ValidSet <- MB_wgroups[-train,]
summary(TrainSet)
summary(ValidSet)


```

```{r create default random forest model on training set}
# Create a Random Forest model with default parameters
model1 <- randomForest(sum_SFN_NIT ~ ., data = TrainSet, importance = TRUE)
model1

R2 <- 1 - (sum((ValidSet$sum_SFN_NIT-predValues)^2)/sum((ValidSet$sum_SFN_NIT-mean(ValidSet$sum_SFN_NIT))^2))
# negative, not a great fit for data

# find number of trees that produced lowest test MSE
which.min(model1$mse)
# 1

# RMSE of best model
sqrt(model1$mse[which.min(model1$mse)])
# 7.1

```

```{r run full model}
model2 <- randomForest(formula = sum_SFN_NIT ~ . , data = MB_wgroups, importance = TRUE)
model2 # not much different than train model

# find number of trees that produced lowest test MSE
which.min(model2$mse)
# 136 trees best

# RMSE of best model
sqrt(model1$mse[which.min(model2$mse)])
# 9.7 is best mse

#plot the test MSE by number of trees
plot(model2)


```

optimize model:
- ntreeTry: The number of trees to build.
- mtryStart: The starting number of predictor variables to consider at each split.
- stepFactor: The factor to increase by until the out-of-bag estimated error stops improving by a certain amount.
- improve: The amount that the out-of-bag error needs to improve by to keep increasing the step factor.
```{r optimize/tune the rf model}
model_tuned <- tuneRF(
               x=MB_wgroups[,-1], #define predictor variables
               y=MB_wgroups$sum_SFN_NIT, #define response variable
               ntreeTry=500,
               mtryStart=2, 
               stepFactor=1.5,
               improve=0.01,
               trace=FALSE #don't show real-time progress
               )
# 27 predictors at each split will give best OOB

```

```{r rerun rf model with new information about best number of predictors}
model3 <- randomForest(formula = sum_SFN_NIT ~ . , data = MB_wgroups, importance = TRUE, mtry = 27)
model3 # not much different than train model

# find number of trees that produced lowest test MSE
which.min(model3$mse)
# 1 tree best, but this changes every time I run it

# RMSE of best model
sqrt(model3$mse[which.min(model3$mse)])
# 5.5 is best mse

# r squared
predValues <- predict(model3, ValidSet)
R2 <- 1 - (sum((ValidSet$sum_SFN_NIT-predValues)^2)/sum((ValidSet$sum_SFN_NIT-mean(ValidSet$sum_SFN_NIT))^2))
# pretty high

#plot the test MSE by number of trees
plot(model3)

# Get variable importance from the model fit
ImpData <- as.data.frame(model3[["importance"]]) %>%
  rename_at('%IncMSE', ~'percIncMSE') %>%
  dplyr::filter(percIncMSE > 0)
ImpData$Var.Names <- row.names(ImpData)

ggplot(ImpData, aes(x=Var.Names, y=`percIncMSE`)) +
  geom_segment( aes(x=Var.Names, xend=Var.Names, y=0, yend=`percIncMSE`), color="skyblue") +
  geom_point(aes(size = IncNodePurity), color="blue", alpha=0.6) +
  theme_light() +
  coord_flip() +
  theme(
    legend.position="bottom",
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank()
  )

# ggsave("importgraph_optimtestmod_AllvarsMBgroupSFN_NIT.png",
#        dpi = 300, height = 4,
#        width = 5)

# another verison of importance plot
varImpPlot(model3, 
           sort=FALSE, 
           main="Variable Importance Plot")


```


## save environment
```{r save env, eval = F}
save.image("brocc_randomforestSFNNIT.RData")

```
