---
title: "broccoli random forest"
author: "Alexandra Alexiev"
date: "2024-11-18"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
# knit options for whole doc. I will add in specific chunks when I don't want this to be the case
knitr::opts_chunk$set(echo = TRUE, # show code
                      eval = TRUE, # run chunks when knitting
                      include = TRUE, # include chunk input in final doc
                      warning = FALSE, # do not show warnings in final doc
                      message = FALSE, # do not show messages from code in final doc
                      collapse = TRUE, # when possible, do put multiple outputs in one block
                      dpi = 300, # fig resolution
                      fig.dim = c(9, 5), # the default figure dimensions
                      out.width = "98%", # the default figure output width
                      out.height = "98%", # the default figure output height 
                      cache = TRUE) # dont rerun chunks that haven't been changed

```


## libraries and directories

```{r libraries}
library(dplyr)
library(tibble)
library(ggplot2)
library(randomForest)
set.seed(3)

```


```{r set working dir, eval=F}
setwd("/Users/alexieva/Documents/Projects/Analysis/broccoli project/02_pubAnalysis/RandomForest/")

load("/Users/alexieva/Documents/Projects/Analysis/broccoli project/02_pubAnalysis/RandomForest/brocc_randomforestSFNNIT.RData")

```

## Input files prep

We want an input file that has the "features" we want to test the importance of vs the samples. The features we want there are the microbial taxa followed by the SFN_NIT data.
```{r file input and clean up}
## input microbiome taxa table
MB_asvtab <- read.delim("/Users/alexieva/Documents/Projects/Analysis/broccoli project/01_explAnalysis/input_data/MBasvtab_brocc_cleanT0.txt") %>%
  rownames_to_column("subject_id")
# str(MB_asvtab)
# MB_asvtab[,2:112] <- log(MB_asvtab[,2:112]+1)

## input SFN_NIT data but combine with asv tab
SFN_NIT_tab <- read.delim("/Users/alexieva/Documents/Projects/Analysis/broccoli project/01_explAnalysis/input_data/Urine_SFN_cumul.txt")

# no NA's?
sum(!complete.cases(MB_asvtab))
sum(!complete.cases(SFN_NIT_tab))
# yes

# for now test with just SFN_NIT_cumulative data
feat_tab <- MB_asvtab %>%
  inner_join(SFN_NIT_tab[c("subject_id", "sum_SFN_NIT")], by = "subject_id") %>%
  column_to_rownames("subject_id")

```

## Split into train and validation set of data
```{r train and validation set split}
# Training Set : Validation Set = 70 : 30 (random)
train <- sample(nrow(feat_tab), 0.7*nrow(feat_tab), replace = FALSE)
TrainSet <- feat_tab[train,]
ValidSet <- feat_tab[-train,]
summary(TrainSet)
summary(ValidSet)


```

## create random forest model to test which taxa are associated with cumulative SFN_NIT
- default uses 500 trees
- regression model

```{r create default random forest model on training set}
# Create a Random Forest model with default parameters
model1 <- randomForest(sum_SFN_NIT ~ ., data = TrainSet, importance = TRUE)
model1
round(importance(model1), 2)

# Get variable importance from the model fit
ImpData <- as.data.frame(importance(model1))
ImpData$Var.Names <- row.names(ImpData)

ggplot(ImpData, aes(x=Var.Names, y=`%IncMSE`)) +
  geom_segment( aes(x=Var.Names, xend=Var.Names, y=0, yend=`%IncMSE`), color="skyblue") +
  geom_point(aes(size = IncNodePurity), color="blue", alpha=0.6) +
  theme_light() +
  coord_flip() +
  theme(
    legend.position="bottom",
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank()
  )

# ggsave("import_graph_trainSFNNIT.png",
#        dpi = 300, height = 13,
#        width = 8)

# calculate RMSE of train data
sqrt(model1$mse[length(model1$mse)]) # want to be low

# calculate RMSE of test data
predValues <- predict(model1, ValidSet)
sqrt(mean((ValidSet$sum_SFN_NIT - predValues)^2))

# ok rmse

R2 <- 1 - (sum((ValidSet$sum_SFN_NIT-predValues)^2)/sum((ValidSet$sum_SFN_NIT-mean(ValidSet$sum_SFN_NIT))^2))
# not a great fit for data

# can't do more mtry's because it's already 37 (and there are 38 observations)
# couldn't make it a better fit by changing ntree

# find number of trees that produced lowest test MSE
which.min(model1$mse) # 1

# RMSE of best model
sqrt(model1$mse[which.min(model1$mse)])

#plot the test MSE by number of trees
plot(model1)

```

```{r run full model}
model2 <- randomForest(formula = sum_SFN_NIT ~ . , data = feat_tab, importance = TRUE)
model2 # not much different than train model

# find number of trees that produced lowest test MSE
which.min(model2$mse)
# 93 trees best, but this changes every time I run it

# RMSE of best model
sqrt(model1$mse[which.min(model2$mse)])
# 31.3 is best mse

#plot the test MSE by number of trees
plot(model2)

# Get variable importance from the model fit
ImpData <- as.data.frame(importance(model2))
ImpData$Var.Names <- row.names(ImpData)

ggplot(ImpData, aes(x=Var.Names, y=`%IncMSE`)) +
  geom_segment( aes(x=Var.Names, xend=Var.Names, y=0, yend=`%IncMSE`), color="skyblue") +
  geom_point(aes(size = IncNodePurity), color="blue", alpha=0.6) +
  theme_light() +
  coord_flip() +
  theme(
    legend.position="bottom",
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank()
  )

# ggsave("import_graph.png",
#        dpi = 300, height = 13,
#        width = 8)


```

optimize model:
- ntreeTry: The number of trees to build.
- mtryStart: The starting number of predictor variables to consider at each split.
- stepFactor: The factor to increase by until the out-of-bag estimated error stops improving by a certain amount.
- improve: The amount that the out-of-bag error needs to improve by to keep increasing the step factor.
```{r optimize/tune the rf model}
model_tuned <- tuneRF(
               x=feat_tab[,-1], #define predictor variables
               y=feat_tab$sum_SFN_NIT, #define response variable
               ntreeTry=500,
               mtryStart=3, 
               stepFactor=1.5,
               improve=0.01,
               trace=FALSE #don't show real-time progress
               )
# 111 predictors at each split will give best OOB

```

```{r rerun rf model with new information about best number of predictors}
model3 <- randomForest(formula = sum_SFN_NIT ~ . , data = feat_tab, importance = TRUE, mtry = 111)
model3 # not much different than train model

# find number of trees that produced lowest test MSE
which.min(model3$mse)
# 105 trees best, but this changes every time I run it

# RMSE of best model
sqrt(model3$mse[which.min(model3$mse)])
# 27.8 is best rmse

#plot the test MSE by number of trees
plot(model3)

# Get variable importance from the model fit
ImpData <- as.data.frame(importance(model3)) %>%
  rename_at('%IncMSE', ~'percIncMSE') %>%
  dplyr::filter(percIncMSE > 0)
ImpData$Var.Names <- row.names(ImpData)

ggplot(ImpData, aes(x=Var.Names, y=`percIncMSE`)) +
  geom_segment( aes(x=Var.Names, xend=Var.Names, y=0, yend=`percIncMSE`), color="skyblue") +
  geom_point(aes(size = IncNodePurity), color="blue", alpha=0.6) +
  theme_light() +
  coord_flip() +
  theme(
    legend.position="bottom",
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank()
  )

# ggsave("import_graph_optimtestmod_SFNNIT.png",
#        dpi = 300, height = 9,
#        width = 8)

# another verison of importance plot
varImpPlot(model3, 
           sort=FALSE, 
           main="Variable Importance Plot")


```

```{r calculate metrics of model fit}
# psuedo r squared
predValues <- predict(model3, ValidSet)
R2 <- 1 - (sum((ValidSet$sum_SFN_NIT-predValues)^2)/sum((ValidSet$sum_SFN_NIT-mean(ValidSet$sum_SFN_NIT))^2))
R2
# pretty ok


```

```{r graph and linear regression of dialister and SFN-NIT}
dialister_NIT_graph <- feat_tab %>%
  dplyr::select(c("Dialister", "sum_SFN_NIT"))

# stats
summary(lm(sum_SFN_NIT ~ Dialister, data = dialister_NIT_graph))
# Adjusted R-squared:  0.1769

# graph
library(ggtext)
dianit_plot <- ggplot(data = dialister_NIT_graph,
                       aes(x = Dialister, y = sum_SFN_NIT)) +
  geom_point() +
  geom_smooth(method = 'lm', color = "black") +
  labs(x = "*Dialister* abundance", y = "cumulative SFN-nitrile in urine") +
  theme(text = element_text(size = 14),
        axis.title.x = ggtext::element_markdown())
dianit_plot
# ggsave("dialisterNIT_lmplot.jpg", dianit_plot,
#        dpi = 300, width = 5, height = 5)


```



## random forest model to test whether the presence or absence of certain taxa is associated with cumulative SFN_NIT

```{r convert asv table into presence absence and add groups}
metadata <- read.delim("../input_data/metadata_wgroups_brocc.txt",
                       header = T, sep = "\t") %>%
  dplyr::select(c("subject_id", "group"))
MB_asvtab_pa <- column_to_rownames(MB_asvtab, "subject_id")
MB_asvtab_pa <- as.data.frame(ifelse(MB_asvtab_pa == 0, "A", "P")) %>%
  rownames_to_column("subject_id") %>%
  inner_join(SFN_NIT_tab[c("subject_id", "sum_SFN_NIT")], by = "subject_id") %>%
  column_to_rownames("subject_id")

## separate into each group
MBasvtab_pa_A <- dplyr::filter(MB_asvtab_pa %>% rownames_to_column("subject_id"), 
                               subject_id %in% metadata$subject_id[metadata$group == "A"])
MBasvtab_pa_B <- dplyr::filter(MB_asvtab_pa %>% rownames_to_column("subject_id"), 
                               subject_id %in% metadata$subject_id[metadata$group == "B"])
MBasvtab_pa_C <- dplyr::filter(MB_asvtab_pa %>% rownames_to_column("subject_id"), 
                               subject_id %in% metadata$subject_id[metadata$group == "C"])


```

```{r split microbiome group data into train and valid set}
# Training Set : Validation Set = 70 : 30 (random)
list <- list(MBasvtab_pa_A, MBasvtab_pa_B, MBasvtab_pa_C)

for (i in list) {
  train <- sample(nrow(i), 0.7*nrow(i), replace = FALSE)
  TrainSet <- i[train,]
  ValidSet <- i[-train,]
  
}

# for whole data
train <- sample(nrow(MB_asvtab_pa), 0.7*nrow(MB_asvtab_pa), replace = FALSE)
TrainSet <- MB_asvtab_pa[train,]
ValidSet <- MB_asvtab_pa[-train,]
summary(TrainSet)
summary(ValidSet)


```

```{r create default random forest model on training set}
# Create a Random Forest model with default parameters
model1 <- randomForest(sum_SFN_NIT ~ ., data = TrainSet, importance = TRUE, type = "class")
model1

R2 <- 1 - (sum((ValidSet$sum_SFN_NIT-predValues)^2)/sum((ValidSet$sum_SFN_NIT-mean(ValidSet$sum_SFN_NIT))^2))
# negative, not a great fit for data

# find number of trees that produced lowest test MSE
which.min(model1$mse)
# 500

# RMSE of best model
sqrt(model1$mse[which.min(model1$mse)])
# 27.4

```

```{r run full model}
model2 <- randomForest(formula = sum_SFN_NIT ~ . , data = MB_asvtab_pa, importance = TRUE)
model2 # not much different than train model

# find number of trees that produced lowest test MSE
which.min(model2$mse)
# 106 trees best

# RMSE of best model
sqrt(model1$mse[which.min(model2$mse)])
# 28 is best mse

#plot the test MSE by number of trees
plot(model2)


```

optimize model:
- ntreeTry: The number of trees to build.
- mtryStart: The starting number of predictor variables to consider at each split.
- stepFactor: The factor to increase by until the out-of-bag estimated error stops improving by a certain amount.
- improve: The amount that the out-of-bag error needs to improve by to keep increasing the step factor.
```{r optimize/tune the rf model}
model_tuned <- tuneRF(
               x=MB_asvtab_pa[,-112], #define predictor variables
               y=MB_asvtab_pa$sum_SFN_NIT, #define response variable
               ntreeTry=500,
               mtryStart=2, 
               stepFactor=1.5,
               improve=0.01,
               trace=FALSE #don't show real-time progress
               )
# 2 predictors at each split will give best OOB

```

```{r rerun rf model with new information about best number of predictors}
model3 <- randomForest(formula = sum_SFN_NIT ~ . , data = MB_asvtab_pa, importance = TRUE, mtry = 2)
model3 # not much different than train model

# find number of trees that produced lowest test MSE
which.min(model3$mse)
# 29 trees best, but this changes every time I run it

# RMSE of best model
sqrt(model3$mse[which.min(model3$mse)])
# 7.8 is best mse

# r squared
predValues <- predict(model3, ValidSet)
R2 <- 1 - (sum((ValidSet$sum_SFN_NIT-predValues)^2)/sum((ValidSet$sum_SFN_NIT-mean(ValidSet$sum_SFN_NIT))^2))
# pretty mid

#plot the test MSE by number of trees
plot(model3)

# Get variable importance from the model fit
ImpData <- as.data.frame(model3[["importance"]]) %>%
  rename_at('%IncMSE', ~'percIncMSE') %>%
  dplyr::filter(percIncMSE > 0)
ImpData$Var.Names <- row.names(ImpData)

ggplot(ImpData, aes(x=Var.Names, y=`percIncMSE`)) +
  geom_segment( aes(x=Var.Names, xend=Var.Names, y=0, yend=`percIncMSE`), color="skyblue") +
  geom_point(aes(size = IncNodePurity), color="blue", alpha=0.6) +
  theme_light() +
  coord_flip() +
  theme(
    legend.position="bottom",
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank()
  )

# ggsave("import_graph_optimtestmod_presabstaxa.png",
#        dpi = 300, height = 8,
#        width = 8)

# another verison of importance plot
varImpPlot(model3, 
           sort=FALSE, 
           main="Variable Importance Plot")


```

## random forest model to test whether microbiome grouping is more important than dietary params in determining SFN_NIT metrics in broccoli consumers
```{r input data with microbiome groups}
MB_wgroups <- read.delim("meta_groups_brocc.txt", header = T, sep = " ") %>%
  dplyr::select(-c("group_numb", "sample", "treatment", "relation")) %>%
  inner_join(SFN_NIT_tab[c("subject_id", "sum_SFN_NIT")], by = "subject_id") %>%
  column_to_rownames("subject_id")


```

```{r split microbiome group data into train and valid set}
# Training Set : Validation Set = 70 : 30 (random)
train <- sample(nrow(MB_wgroups), 0.7*nrow(MB_wgroups), replace = FALSE)
TrainSet <- MB_wgroups[train,]
ValidSet <- MB_wgroups[-train,]
summary(TrainSet)
summary(ValidSet)


```

```{r create default random forest model on training set}
# Create a Random Forest model with default parameters
model1 <- randomForest(sum_SFN_NIT ~ ., data = TrainSet, importance = TRUE)
model1

R2 <- 1 - (sum((ValidSet$sum_SFN_NIT-predValues)^2)/sum((ValidSet$sum_SFN_NIT-mean(ValidSet$sum_SFN_NIT))^2))
# negative, not a great fit for data

# find number of trees that produced lowest test MSE
which.min(model1$mse)
# 1

# RMSE of best model
sqrt(model1$mse[which.min(model1$mse)])
# 7.1

```

```{r run full model}
model2 <- randomForest(formula = sum_SFN_NIT ~ . , data = MB_wgroups, importance = TRUE)
model2 # not much different than train model

# find number of trees that produced lowest test MSE
which.min(model2$mse)
# 136 trees best

# RMSE of best model
sqrt(model1$mse[which.min(model2$mse)])
# 9.7 is best mse

#plot the test MSE by number of trees
plot(model2)


```

optimize model:
- ntreeTry: The number of trees to build.
- mtryStart: The starting number of predictor variables to consider at each split.
- stepFactor: The factor to increase by until the out-of-bag estimated error stops improving by a certain amount.
- improve: The amount that the out-of-bag error needs to improve by to keep increasing the step factor.
```{r optimize/tune the rf model}
model_tuned <- tuneRF(
               x=MB_wgroups[,-1], #define predictor variables
               y=MB_wgroups$sum_SFN_NIT, #define response variable
               ntreeTry=500,
               mtryStart=2, 
               stepFactor=1.5,
               improve=0.01,
               trace=FALSE #don't show real-time progress
               )
# 27 predictors at each split will give best OOB

```

```{r rerun rf model with new information about best number of predictors}
model3 <- randomForest(formula = sum_SFN_NIT ~ . , data = MB_wgroups, importance = TRUE, mtry = 27)
model3 # not much different than train model

# find number of trees that produced lowest test MSE
which.min(model3$mse)
# 1 tree best, but this changes every time I run it

# RMSE of best model
sqrt(model3$mse[which.min(model3$mse)])
# 5.5 is best mse

# r squared
predValues <- predict(model3, ValidSet)
R2 <- 1 - (sum((ValidSet$sum_SFN_NIT-predValues)^2)/sum((ValidSet$sum_SFN_NIT-mean(ValidSet$sum_SFN_NIT))^2))
# pretty high

#plot the test MSE by number of trees
plot(model3)

# Get variable importance from the model fit
ImpData <- as.data.frame(model3[["importance"]]) %>%
  rename_at('%IncMSE', ~'percIncMSE') %>%
  dplyr::filter(percIncMSE > 0)
ImpData$Var.Names <- row.names(ImpData)

ggplot(ImpData, aes(x=Var.Names, y=`percIncMSE`)) +
  geom_segment( aes(x=Var.Names, xend=Var.Names, y=0, yend=`percIncMSE`), color="skyblue") +
  geom_point(aes(size = IncNodePurity), color="blue", alpha=0.6) +
  theme_light() +
  coord_flip() +
  theme(
    legend.position="bottom",
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank()
  )

# ggsave("importgraph_optimtestmod_AllvarsMBgroupSFN_NIT.png",
#        dpi = 300, height = 4,
#        width = 5)

# another verison of importance plot
varImpPlot(model3, 
           sort=FALSE, 
           main="Variable Importance Plot")


```


## save environment
```{r save env, eval = F}
save.image("brocc_randomforestSFNNIT.RData")

```
