---
title: "brocc mediation on time 0 clusters"
author: "Alexandra Alexiev"
date: "2024-07-09"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
# knit options for whole doc. I will add in specific chunks when I don't want this to be the case
knitr::opts_chunk$set(echo = TRUE, # show code
                      eval = TRUE, # run chunks when knitting
                      include = TRUE, # include chunk input in final doc
                      warning = FALSE, # do not show warnings in final doc
                      message = FALSE, # do not show messages from code in final doc
                      collapse = TRUE, # when possible, do put multiple outputs in one block
                      dpi = 300, # fig resolution
                      fig.dim = c(9, 5), # the default figure dimensions
                      out.width = "98%", # the default figure output width
                      out.height = "98%", # the default figure output height 
                      cache = TRUE) # dont rerun chunks that haven't been changed

```


## libraries and directories

```{r libraries}
library(tidyverse)
library(ggplot2)
library(vegan)
library(mediation)
library(ggfortify)
library(ggpubr)
library(tweedie)
library(statmod)
library(AICcmodavg)
library(RColorBrewer)
set.seed(3)

```


```{r set working dir, eval=F}
setwd("/Users/alexieva/Documents/Projects/Analysis/broccoli project/02_pubAnalysis/Prediction2_mediation")

load("/Users/alexieva/Documents/Projects/Analysis/broccoli project/02_pubAnalysis/Prediction2_mediation/brocc_clustermediation.RData")

```

## Input data
```{r input microbiome asv table, urine SFN data, and diet data}
## microbiome data
# taxa table
MB_taxtab <- read.csv("/Users/alexieva/Documents/Projects/Analysis/broccoli project/00_data/Microbiome_Flat/TaxTable_BSS.csv",
                      header = TRUE) 

# metadata for microbiome stuff
metadata <- read.csv("/Users/alexieva/Documents/Projects/Analysis/broccoli project/00_data/Microbiome_Flat/SampleData_BSS.csv",
                      header = TRUE) %>%
  dplyr::filter(time == 0) %>% # for now just Time 0 data
  mutate(broc_consum = case_when(veg == "broc" ~ "1",
                                 veg == "alf" ~ "0"))

# ASV table (ASV reads per sample)
MB_asvtab <- read.csv("/Users/alexieva/Documents/Projects/Analysis/broccoli project/00_data/Microbiome_Flat/ASVTable_BSS.csv",
                      header = TRUE) %>%
  dplyr::filter(sample %in% metadata$sample) # just time 0 data

# clean up subject ID names so they will match the metadata
MB_asvtab$sample <- sub("_[^_]+_[^_]+$", "", MB_asvtab$sample)

# rename the ASVs by their taxonomy (genus level)
MB_asvtab <-as.data.frame(t(column_to_rownames(MB_asvtab, "sample")))
MB_asvtab$Genus[which(rownames(MB_asvtab) == MB_taxtab$ASV)] <- MB_taxtab$Genus
MB_asvtab <- MB_asvtab %>%
  group_by(Genus) %>%
  summarise(across(where(is.numeric), sum)) %>%
  column_to_rownames("Genus") %>%
  t() %>%
  as.data.frame()


```

```{r SFN data}
## SFN data
# filtering just timepoint 24 hours, as per Laura Beaver's observations that this was the peak SFN and SFN-NIT output in urine
# filtering just SFN and SFN-NIT because those were the significantly correlated ones to broccoli consumption in Dr. Beaver's previous analysis
## SFN data

Urine_SFN <- read.csv("/Users/alexieva/Documents/Projects/Analysis/broccoli project/00_data/SFN_Data/Urine_R_Format.csv",
                      header = TRUE) %>%
    group_by(subject_id) %>%
  summarise(sum_SFN = sum(SFN),
            sum_SFN_Cys = sum(SFN_Cys),
            sum_SFN_NAC = sum(SFN_NAC),
            sum_SFN_CG = sum(SFN_CG),
            sum_SFN_GSH = sum(SFN_GSH),
            sum_SFN_NIT = sum(SFN_NIT),
            sum_SFN_Tot = sum(SFN_Tot))

# make a mega metadata file
mega_meta <- inner_join(metadata, Urine_SFN, by = "subject_id")
mega_meta <- mega_meta[order(mega_meta$subject_id),] # this also takes out the alfalfa eaters

```

```{r diet data}
## diet data
# make a list of the "significant" diet factors we are interested in 
diet_compnts <- c("diet_Chol_mg", "diet_SatFat_g", "diet_Alc_g", "diet_TotSolFib_g", 
                  "diet_MonSac_g", "diet_Disacc_g", "diet_BetaCaro_mcg", "diet_Vit_B1_mg",
                  "diet_Vit_B6_mg", "diet_Fol.DFE_mcg_DFE", "diet_Mang_mg", "diet_Copp_mg", 
                  "diet_TotFib_g", "diet_OCarb_g", "diet_Iron_mg", "diet_TotInsolFib_g")
diet <- read.csv("/Users/alexieva/Documents/Projects/Analysis/broccoli project/00_data/Diet/New_Diet_Small.csv", header = TRUE) %>%
  dplyr::select(c("subject_id", all_of(diet_compnts)))

```

```{r cluster beta diversity}
bray_nmds <- metaMDS(as.matrix(MB_asvtab), distance = "bray", 
                     k = 3, trymax = 100)
bray_nmds$stress # is less than 0.2 so good stress

ordiplot(bray_nmds) # no clear groupings/clusters

d <- vegdist(MB_asvtab, method = "bray") # calc bray curtis dist
hc <- hclust(d, method = "complete") # cluster dendrogram
plot(hc) # looks like five groups could be made at a macro level

groups <- as.data.frame(cutree(hc, 5)) # assign 1-5 group IDs to each sample ID based on hclust
names(groups) <- "group"
# combine with metadata for graphing
meta_groups <- groups %>%
  rownames_to_column("subject_id") %>%
  inner_join(metadata, by = "subject_id")

# is diet related to BC dissimilarity?
adonis2(d ~ diet_TotFib_g, data = diet, method = "bray")
# yes
adonis2(d ~ diet_TotSolFib_g, data = diet, method = "bray")
# yes

# loop through all
out <- data.frame(matrix(ncol = 3, nrow = 0))
colnames(out) <- c("name", "r2", "p")
j <- 1
for (i in 1:length(diet_compnts)) {
  diet2 <- diet %>% dplyr::mutate(comp = diet[[diet_compnts[i]]])
  perm <- adonis2(d ~ comp, data = diet2, method = "bray")
  if (perm$`Pr(>F)`[1] <= 0.05) {
    out[j,1] <- diet_compnts[i]
    out[j,2] <- perm$R2[1]
    out[j,3] <- perm$`Pr(>F)`[1]
    j <- j + 1
  }
}

out


```

## a more precise way of determining clusters and number of clusters
K means with silhouette score
```{r k means}
library(cluster) 
library(factoextra)

## with alfalfa eaters
fviz_nbclust(MB_asvtab, kmeans, method = 'silhouette')

# 2 is actualy the optimal clustering for the ASV matrix

## without alfalfa eaters (only brocc)
fviz_nbclust(MB_asvtab_filt, kmeans, method = 'silhouette')

# 3 is actualy the optimal clustering for the ASV matrix


```


## Mediation with cluster

```{r combine above separate data constructs into one with all info necessary for model}
# match data based on subject id
# filter only broccoli eaters now
groups <- kmeans(MB_asvtab_filt, centers = 3, nstart = 25)
groups_r <- as.data.frame(groups$cluster) %>%
  rownames_to_column("subject_id")
names(groups_r)[names(groups_r) == 'groups$cluster'] <- 'group'
meta_filt <- metadata[which(metadata$veg == "broc"),]

med_input <- MB_asvtab %>%
  rownames_to_column("subject_id") %>%
  dplyr::filter(subject_id %in% meta_filt$subject_id) %>%
  inner_join(Urine_SFN, by = "subject_id") %>%
  inner_join(diet, by = "subject_id") %>%
  inner_join(groups_r, by = "subject_id")

med_input$group[med_input$group == 1] <- "A"
med_input$group[med_input$group == 2] <- "B"
med_input$group[med_input$group == 3] <- "C"


```

```{r save input files for other analyses}
# metadata table with groups including alfalfa eaters
groups_alf <- kmeans(MB_asvtab, centers = 2, nstart = 25)
groups_r_alf <- as.data.frame(groups_alf$cluster) %>%
  rownames_to_column("subject_id")
meta_groups <- metadata %>%
  inner_join(groups_r_alf, by = "subject_id")
meta_groups$group[meta_groups$group == 1] <- "A"
meta_groups$group[meta_groups$group == 2] <- "B"
# write.table(meta_groups,
#             "/Users/alexieva/Documents/Projects/Analysis/broccoli project/01_explAnalysis/brocc_mediation_Rfiles/inputs/meta_groups_alfbrocc.txt",
#             sep = "\t")

# metadata table with groups 
meta_groups_brocc <- meta_filt %>%
  inner_join(groups_r, by = "subject_id")
meta_groups_brocc$group[meta_groups_brocc$group == 1] <- "A"
meta_groups_brocc$group[meta_groups_brocc$group == 2] <- "B"
meta_groups_brocc$group[meta_groups_brocc$group == 3] <- "C"
# write.table(meta_groups_brocc,
#             "/Users/alexieva/Documents/Projects/Analysis/broccoli project/01_explAnalysis/brocc_mediation_Rfiles/inputs/meta_groups_brocc.txt",
#             sep = "\t")

```


```{r revise groups, eval=F}
med_input <- med_input %>%
  mutate(group_numb = case_when(group == "A" ~ 1,
         group == "B" ~ 2,
         group == "C" ~ 3))


```

```{r enterotypes?}
## see if there are enterotypes with these groupings
table(med_input$group)
samp_numbs <- count(med_input, group)
# group C is lowest but ok for now

## clean up and calculate sums and proportions
med_input_taxgroups_med <- med_input %>% 
  column_to_rownames("subject_id") %>%
  dplyr::select(-c("group_numb", contains("sum"), contains("diet"))) %>%
  mutate(Other = dplyr::select(., -c("Bacteroides", "Prevotella_9", contains("Ruminococcus"), "group")) %>% rowSums()) %>%
  mutate(Rumino_sum = dplyr::select(., contains("Ruminococcus")) %>% rowSums())

med_input_taxsummed <- subset(med_input_taxgroups_med, select = c(Bacteroides, Prevotella_9, Rumino_sum, Other, group))

med_input_taxgroups <- med_input_taxsummed %>%
  group_by(group) %>%
  summarise(across(everything(), sum, na.rm = TRUE)) %>%
  column_to_rownames("group")

med_prop_taxgroups <- prop.table(as.matrix(med_input_taxgroups), margin = 1) %>%
  as.data.frame() %>%
  rownames_to_column("group") %>%
  rename(Ruminoccocus = Rumino_sum) %>%
  gather("taxa", "proportion", -group)

# reorder factors
med_prop_taxgroups$taxa <- factor(med_prop_taxgroups$taxa, 
                                    levels=c('Bacteroides', 'Prevotella_9', 
                                             'Ruminoccocus', 'Other'))
# set pallete
pal <- c(brewer.pal(n = 3, name = "Dark2"), "grey")

# graph
ggplot(med_prop_taxgroups, aes(fill = taxa, y = proportion, x = group)) + 
  geom_bar(position = 'stack', stat = 'identity') +
  scale_fill_manual(values = pal)


# ggsave("props_enterotype.png",
#        dpi = 300, height = 5, width = 6)


```

- check if e.g., total fiber is associated with groupings
```{r model Y ~ M, cluster group ~ diet}
# plot of fiber in each group
ggplot(data = med_input, aes(x = group,
                             y = diet_TotFib_g)) +
  geom_boxplot()
# density plot of total fiber
ggplot(data = med_input, aes(x = diet_TotFib_g)) +
  geom_density()

# plot of SFN in each group
ggplot(data = med_input, aes(x = group,
                             y = sum_SFN)) +
  geom_boxplot()


model.M <- lm(diet_TotFib_g ~ group, data = med_input)
summary(model.M)

model.M <- lm(sum_SFN ~ group, data = med_input)
summary(model.M)

```

- trouble shoot model Y to choose optimal model
```{r model Y ~ X + M, SFN ~ cluster group + diet}
model.Y <- lm(log(sum_SFN+1) ~ group + diet_TotFib_g, data = med_input)
summary(model.Y)

model.Y.1 <- glm(log(sum_SFN+1) ~ group + diet_TotFib_g, data = med_input,
                 family = Gamma)
summary(model.Y.1)

model.Y.2 <- glm(log(sum_SFN+1) ~ group + diet_TotFib_g, med_input, 
                   family = inverse.gaussian(link = "log"))
summary(model.Y.2)

AIC(model.Y, model.Y.1, model.Y.2) # lm is the one


```

```{r test a mediation}
med_input_l <- med_input %>%
        dplyr::mutate(sfn_trans = log(med_input[[SFN_list[1]]] + 1),
                      quarts = ntile(med_input[[diet_compnts[1]]], 4)) %>%
  column_to_rownames("subject_id")
med_input_l$group <- as.factor(med_input_l$group)

# set models
model.M <- lm(diet_Chol_mg ~ group,
                    med_input_l)
summary(model.M)
model.Y <- lm(sfn_trans ~ group + diet_Chol_mg,
                    med_input_l)
summary(model.Y)
      
# mediate
results <- mediate(model.M, model.Y, 
                         treat = "group", 
                         mediator = "diet_Chol_mg",
                         control.value = "A", treat.value = "B", # lowest and highest quartiles
                         boot = TRUE, sims = 500)

summary(results)


```

# test model 0: which groups vary with SFN?
```{r model 0 test}
# loop through all
out <- data.frame(matrix(ncol = 1, nrow = 0))
colnames(out) <- c("name")
long <- c()
j <- 1
for (i in 1:length(SFN_list)) {
  input <- data.frame("sfn" = med_input[[SFN_list[i]]], 
                      "group" = med_input$group,
                      "subject_id" = med_input$subject_id)
  lm <- summary(lm(sfn ~ group, data = input))
  long[[SFN_list[i]]] <- as.data.frame(lm$coefficients)
  if (lm$coefficients[2,4] <= 0.05 | lm$coefficients[3,4] <= 0.05) {
    out[j,1] <- SFN_list[i]
    j <- j + 1
  }
}

out
# ok so none of these vary over group, but sfn_nit is the closest to being significant
# also these are lms so not pairwise comparisons

```


```{r mediation loop}
# Use list_sfn and diet_compnts as the list of dietary components
# make an output file
med_res <- c()
pvals_med <- c()
exp_res_med <- c()
SFN_list <- c("sum_SFN", "sum_SFN_NIT", "sum_SFN_Tot", "sum_SFN_Cys",
              "sum_SFN_NAC", "sum_SFN_CG", "sum_SFN_GSH", "sum_SFN_NIT")
# list_combos <- data.frame(sfns = c(rep("sum_SFN", nrow(out)),
#                                    rep("sum_SFN_NIT", nrow(out)),
#                                    rep("sum_SFN_Tot", nrow(out))),
#                           diet = out$name)
# list_groups <- unique(med_input$group)

for (a in 1:length(diet_compnts)) {
    for (f in 1:length(SFN_list)) {
      tryCatch({
      # add quartiles for each dietary component that will be used
      # transform ASV abundance and transform SFN abundance
      med_input_l <- med_input %>%
        dplyr::mutate(sfn_trans = log(med_input[[SFN_list[f]]] + 1),
                      quarts = med_input[[diet_compnts[a]]]) %>%
                      # quarts = ntile(med_input[[diet_compnts[a]]], 4)) %>%
  column_to_rownames("subject_id")
      med_input_l$group <- as.factor(med_input_l$group)
      
      # set models
      model.M <- lm(quarts ~ group,
                    med_input_l)
      model.Y <- lm(sfn_trans ~ quarts + group,
                    med_input_l)
      
      # mediate
      resultsAB <- mediate(model.M, model.Y, 
                         treat = "group", 
                         mediator = "quarts",
                         control.value = "A", treat.value = "B", # lowest and highest quartiles
                         boot = TRUE, sims = 500)
      resultsAC <- mediate(model.M, model.Y,
                         treat = "group",
                         mediator = "quarts",
                         control.value = "A", treat.value = "C", # lowest and highest quartiles
                         boot = TRUE, sims = 500)
      resultsBC <- mediate(model.M, model.Y,
                         treat = "group",
                         mediator = "quarts",
                         control.value = "B", treat.value = "C", # lowest and highest quartiles
                         boot = TRUE, sims = 500)
      
      # save results
      exp_res_med[[paste(SFN_list[f], diet_compnts[a], "_AB")]] <- resultsAB
      exp_res_med[[paste(SFN_list[f], diet_compnts[a], "_BC")]] <- resultsBC
      exp_res_med[[paste(SFN_list[f], diet_compnts[a], "_AC")]] <- resultsBC
      modsumAB <- summary(resultsAB)
      modsumBC <- summary(resultsBC)
      modsumAC <- summary(resultsAC)
      pvals_med[[paste(SFN_list[f], diet_compnts[a], "_AB")]] <- list(ACME_pval_AB = modsumAB$d.avg.p,
                                                               ADE_pval_AB = modsumAB$z.avg.p,
                                                               prop_mediated_pval_AB = modsumAB$n.avg.p)
      pvals_med[[paste(SFN_list[f], diet_compnts[a], "_BC")]] <- list(ACME_pval_BC = modsumBC$d.avg.p,
                                                               ADE_pval_BC = modsumBC$z.avg.p,
                                                               prop_mediated_pval_BC = modsumBC$n.avg.p)
      pvals_med[[paste(SFN_list[f], diet_compnts[a], "_AC")]] <- list(ACME_pval_AC = modsumAC$d.avg.p,
                                                               ADE_pval_AC = modsumAC$z.avg.p,
                                                               prop_mediated_pval_AC = modsumAC$n.avg.p)},
      error = function(e){cat("ERROR :",conditionMessage(e), "\n")})
      
      if (modsumAB$d.avg.p <= 0.2) {
        coeff_tab <- list(ACME_est = modsumAB$d.avg,
                          ACME_pval = modsumAB$d.avg.p,
                          ADE_est = modsumAB$z.avg,
                          ADE_pval = modsumAB$z.avg.p,
                          prop_mediated_est = modsumAB$n.avg,
                          prop_mediated_pval = modsumAB$n.avg.p)
        
        med_res[[paste(SFN_list[f], diet_compnts[a], "_AB")]] <- coeff_tab
        
        if (modsumBC$d.avg.p <= 0.2) {
          coeff_tab <- list(ACME_est = modsumBC$d.avg,
                          ACME_pval = modsumBC$d.avg.p,
                          ADE_est = modsumBC$z.avg,
                          ADE_pval = modsumBC$z.avg.p,
                          prop_mediated_est = modsumBC$n.avg,
                          prop_mediated_pval = modsumBC$n.avg.p)
        
        med_res[[paste(SFN_list[f], diet_compnts[a], "_BC")]] <- coeff_tab
        
        if (modsumAC$d.avg.p <= 0.2) {
          coeff_tab <- list(ACME_est = modsumAC$d.avg,
                          ACME_pval = modsumAC$d.avg.p,
                          ADE_est = modsumAC$z.avg,
                          ADE_pval = modsumAC$z.avg.p,
                          prop_mediated_est = modsumAC$n.avg,
                          prop_mediated_pval = modsumAC$n.avg.p)
        
        med_res[[paste(SFN_list[f], diet_compnts[a], "_AC")]] <- coeff_tab
        
        }
        }
        
      } else { print(paste(SFN_list[f], ",",
                           diet_compnts[a], "are NOT significant"))}
      
      # counters
      f <- f + 1
      }
    a <- a + 1
    }



```

## Examine results
```{r which SFN-taxa-diet combos are potential mediations}
signif <- do.call(rbind.data.frame, pvals_med[names(med_res)]) %>%
  rownames_to_column("names(med_res)")
signif

# write.table(signif, "../outputs/signif_meds_pred2.txt")

# after adding insoluble fiber, same answer

```

one significant association, indirect effect

## graph the results
```{r graphing one result}
## sum_SFN_NIT diet_OCarb_g _AB

# select and plot mediation result for this combo
med_res$`sum_SFN_NIT diet_OCarb_g _AB`
# it's not great beyond the ACME
plot(summary(exp_res_med$`sum_SFN_NIT diet_OCarb_g _AB`))

# select these variables for plotting
regrplot <- med_input %>% dplyr::select(group, sum_SFN_NIT, diet_OCarb_g)

# graph model.M
model.M <- lm(diet_OCarb_g ~ group, med_input)
summary(model.M)
# group B almost significant

modM <- ggplot(data = regrplot, aes(x = group, 
                                    y = diet_OCarb_g)) +
  geom_boxplot(aes(group = group)) +
  labs(title = "Microbiome group ~ Other Carbs", y = "Other Carbs (g)")
modM

# graph model.Y
model.Y <- lm(log(sum_SFN_NIT+1) ~ diet_OCarb_g + group, med_input)
summary(model.Y)

modY_ASV <- ggplot(data = regrplot, aes(x = group, 
                                        y = log(sum_SFN_NIT+1))) +
  geom_boxplot(aes(group = group)) +
  ggtitle("SFN NIT ~ Microbiome group")

modY_fib <- ggplot(data = regrplot, aes(x = diet_OCarb_g, y = log(sum_SFN_NIT+1), color = group)) +
  geom_point() +
  stat_smooth(method = "lm", se = TRUE) +
  labs(title = "SFN NIT ~ Other Carbs",
       x = "Other Carbs (g)")

ggarrange(modM, modY_ASV, modY_fib,
          nrow = 1, ncol = 3)

# ggsave("clustermediation.png",
#        height = 5, width = 15, dpi = 300)

```


## save image chunk 
```{r save env, eval = F}
save.image("/Users/alexieva/Documents/Projects/Analysis/broccoli project/01_explAnalysis/brocc_mediation_Rfiles/cluster_method/brocc_clustermediation.RData")

```
