---
title: "DataFileCleanupBrocc"
author: "Alexandra Alexiev"
date: "2024-08-13"
output: pdf_document
---

```{r setup, include=FALSE}
# knit options for whole doc. I will add in specific chunks when I don't want this to be the case
knitr::opts_chunk$set(echo = TRUE, # show code
                      eval = TRUE, # run chunks when knitting
                      include = TRUE, # include chunk input in final doc
                      warning = FALSE, # do not show warnings in final doc
                      message = FALSE, # do not show messages from code in final doc
                      collapse = TRUE, # when possible, do put multiple outputs in one block
                      dpi = 300, # fig resolution
                      fig.dim = c(9, 5), # the default figure dimensions
                      out.width = "98%", # the default figure output width
                      out.height = "98%", # the default figure output height 
                      cache = TRUE) # dont rerun chunks that haven't been changed

```

## libraries and directories

```{r libraries}
library(dplyr)
library(tibble)
library(tidyr)
set.seed(3)

```

```{r set working dir, eval=F}
setwd("/Users/alexieva/Documents/Projects/Analysis/broccoli project/01_explAnalysis/")

input_folder <- "/Users/alexieva/Documents/Projects/Analysis/broccoli project/01_explAnalysis/input_data/"

```


## Microbiome data
- time 24 hr microbiome data filtered
- aggregate read counts by Genus

```{r MB data}
# taxa table
MB_taxtab <- read.csv("/Users/alexieva/Documents/Projects/Analysis/broccoli project/00_data/Microbiome_Flat/TaxTable_BSS.csv",
                      header = TRUE) 

# metadata for microbiome stuff
metadata <- read.csv("/Users/alexieva/Documents/Projects/Analysis/broccoli project/00_data/Microbiome_Flat/SampleData_BSS.csv",
                      header = TRUE) %>%
  dplyr::filter(time == 24, veg == "broc") # for now just Time 24 data and only broccoli consumers

# ASV table (ASV reads per sample)
MB_asvtab <- read.csv("/Users/alexieva/Documents/Projects/Analysis/broccoli project/00_data/Microbiome_Flat/ASVTable_BSS.csv",
                      header = TRUE) %>%
  dplyr::filter(sample %in% metadata$sample) # just time 0 data

# clean up subject ID names so they will match the metadata
MB_asvtab$sample <- sub("_[^_]+_[^_]+$", "", MB_asvtab$sample)

# rename the ASVs by their taxonomy (genus level)
MB_asvtab <-as.data.frame(t(column_to_rownames(MB_asvtab, "sample")))
MB_asvtab$Genus[which(rownames(MB_asvtab) == MB_taxtab$ASV)] <- MB_taxtab$Genus
MB_asvtab <- MB_asvtab %>%
  group_by(Genus) %>%
  summarise(across(where(is.numeric), sum)) %>%
  t() %>%
  as.data.frame()
colnames(MB_asvtab) <- MB_asvtab[1,]
MB_asvtab <- MB_asvtab[-1,]
MB_asvtab <- MB_asvtab[order(rownames(MB_asvtab)),]


```

```{r save MB objects}
write.table(MB_asvtab, paste0(input_folder, "MB_asvtab_24hr.txt"))
write.table(metadata, paste0(input_folder, "brocc_metadata_24hr.txt"))

```



## SFN data
- from Urine
- 24 hr data, Laura suggested this is best time point to use as it is the peak
- SFN and SFN-NIT most correlated with broccoli consumption according to Laura
```{r}
Urine_SFN <- read.table("/Users/alexieva/Documents/Projects/Analysis/broccoli project/01_explAnalysis/Urine_tab_rawlong.txt", header = TRUE) %>%
  dplyr::filter(time == 24 & SFN_metabolites == "SFN" | time == 24 & SFN_metabolites == "SFN_NIT") %>%
  spread(key = "SFN_metabolites", value = "quantity") %>%
  dplyr::select(-time)


```

```{r save MB objects}
write.table(Urine_SFN, paste0(input_folder, "Urine_SFN_rawlong.txt"))


```

## Diet composition data
- based on Yanni's previous research
- taken in the weeks leading up to Time 0 experiment
```{r diet comp}
diet_compnts <- c("diet_Chol_mg", "diet_SatFat_g", "diet_Alc_g", "diet_TotSolFib_g", 
                  "diet_MonSac_g", "diet_Disacc_g", "diet_BetaCaro_mcg", "diet_Vit_B1_mg",
                  "diet_Vit_B6_mg", "diet_Fol.DFE_mcg_DFE", "diet_Mang_mg", "diet_Copp_mg", 
                  "diet_TotFib_g", "diet_OCarb_g", "diet_Iron_mg")

diet <- read.csv("/Users/alexieva/Documents/Projects/Analysis/broccoli project/00_data/Diet/Diet_Small.csv", header = TRUE) %>%
  dplyr::select(c("subject_id", all_of(diet_compnts)))


```

```{r save MB objects}
write.table(diet, paste0(input_folder, "diet_filtered.txt"))


```


